---
title: "Trajectory checks: stress-testing the AEM–RDA regime workflow"
subtitle: "Pilot diagnostics for the EcoIndicators Synthesis manuscript (13 sanctuaries × 7 EOVs)"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
    code-tools: true
execute:
  echo: true
  message: false
  warning: false
  error: true
editor_options:
  chunk_output_type: console
---

<!--
PROMPT to Claude CODE 2026-10-03 (abridged):

Interrogate the following notions in the EcoIndicators Synthesis paper: (1) can we illustrate the
ordination techniques against known past events in the Sanctuaries (cold snap in Florida Keys, anoxic
event in Flower Garden Banks, heatwave in California)? (2) is the assumption that an alternating
dominant axis is a sign of an eco-paradigm shift a safe one? (3) can we draw reasonable conclusions from
seasonal time binning, or is there another way to capture seasonality without predefined segmentation?
(4) what other indicator data should we include? Any code that you write and keep for analysis should
be committed to one of the data repos.
-->

These are **independent pilot checks**, not the manuscript's results. They re-implement the workflow
shown in the 2026-04-20 working-group deck (monthly anomalies → z-scores → seasonal subsets →
asymmetric eigenvector maps (AEM) → forward selection → RDA → fitted-axis "temporal control chart")
from the same extracts ([copernicus](copernicus.html), [erddap](erddap.html)) and then ask how the
conclusions hold up when a choice is varied or a null model is applied.

```{r}
#| label: setup

librarian::shelf(
  adespatial, dplyr, ecp, ggplot2, glue, here, knitr, lubridate, parallel,
  patchwork, purrr, readr, scales, stringr, tibble, tidyr, vegan,
  quiet = T)
options(readr.show_col_types = F)
set.seed(42)

# paths ----
copernicus_csv <- here("data/copernicus.csv")
erddap_csv     <- here("data/erddap.csv")
dir_chk        <- here("data/trajectory_checks")
dir_fig        <- here("figures/trajectory_checks")
deck_csv       <- file.path(dir_chk, "deck_2026-04-20_r2adj.csv")
null_csv       <- file.path(dir_chk, "null_52.csv")
full_csv       <- file.path(dir_chk, "full_series_aem.csv")
cp_csv         <- file.path(dir_chk, "changepoints.csv")
dir.create(dir_chk, showWarnings = F, recursive = T)
dir.create(dir_fig, showWarnings = F, recursive = T)

# constants ----
N_SUR   <- 200                       # surrogates per analysis
N_CORE  <- max(1, detectCores() - 2)
VARS    <- c("tempSurf", "tempBott", "mixThick", "salinity", "precip", "CHL", "PP")
SEASONS <- c("winter", "spring", "summer", "fall")
# sanctuaries, ordered pacific (n -> s, then islands) to gulf and atlantic (s -> n)
NMS_ORD <- c(
  "OCNMS", "GFNMS", "CBNMS", "MBNMS", "CPNMS", "CINMS", "HIHWNMS", "NMSAS",
  "FGBNMS", "FKNMS", "GRNMS", "MNMS", "SBNMS")
VAR_LUT <- c(
  "thetao_depth=0.49402499" = "tempSurf",
  "bottomT"                 = "tempBott",
  "mlotst"                  = "mixThick",
  "so_depth=0.49402499"     = "salinity",
  "precipitation"           = "precip",
  "CHL"                     = "CHL",
  "PP"                      = "PP")

# plot theme: recessive chrome, thin marks ----
PAL <- list(
  surface = "#fcfcfb", ink = "#0b0b0b", ink2 = "#52514e", muted = "#898781",
  grid    = "#e1e0d9", axis = "#c3c2b7", blue = "#2a78d6", orange = "#eb6834")
PAL_DIV  <- c("#104281", "#2a78d6", "#9ec5f4", "#f0efec", "#f3b3ad", "#e34948", "#8f1d1c")
PAL_SEQ  <- c("#f0efec", "#cde2fb", "#86b6ef", "#3987e5", "#1c5cab", "#0d366b")
PAL_BAND <- c("#86b6ef", "#3987e5", "#1c5cab", "#0d366b")   # ordinal, short -> long time scale

theme_chk <- function(base_size = 10) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.background    = element_rect(fill = PAL$surface, color = NA),
      panel.background   = element_rect(fill = PAL$surface, color = NA),
      panel.grid.major   = element_line(color = PAL$grid, linewidth = 0.25),
      panel.grid.minor   = element_blank(),
      axis.text          = element_text(color = PAL$muted),
      axis.title         = element_text(color = PAL$ink2),
      plot.title         = element_text(color = PAL$ink, face = "bold", size = rel(1.1)),
      plot.subtitle      = element_text(color = PAL$ink2),
      plot.caption       = element_text(color = PAL$muted, hjust = 0),
      plot.title.position = "plot",
      strip.text         = element_text(color = PAL$ink, face = "bold", hjust = 0),
      legend.position    = "top",
      legend.justification = "left",
      legend.title       = element_text(color = PAL$ink2),
      legend.text        = element_text(color = PAL$ink2))
}
save_fig <- function(p, name, width = 9, height = 5) {
  ggsave(file.path(dir_fig, name), p, width = width, height = height, dpi = 200, bg = PAL$surface)
  p
}
```

## Data

Sanctuary-wide monthly means of seven variables for thirteen sanctuaries. Where the multi-year (`my`)
and interim (`myint`) GLORYS12 streams overlap, the multi-year value is kept.

```{r}
#| label: data

d_wide <- bind_rows(read_csv(copernicus_csv), read_csv(erddap_csv)) |>
  filter(
    nms  %in% NMS_ORD,
    var  %in% names(VAR_LUT),
    stat == "mean") |>
  mutate(
    date   = as_date(time),
    var    = recode(var, !!!VAR_LUT),
    is_int = str_detect(dataset, "myint")) |>
  arrange(nms, var, date, is_int) |>
  distinct(nms, var, date, .keep_all = T) |>
  select(nms, date, var, value) |>
  pivot_wider(names_from = var, values_from = value) |>
  arrange(nms, date)

d_wide |>
  pivot_longer(all_of(VARS), names_to = "var") |>
  drop_na(value) |>
  group_by(var) |>
  summarize(
    first = min(date), last = max(date), sanctuaries = n_distinct(nms),
    .groups = "drop") |>
  kable(caption = "Coverage of the seven analysis variables")
```

## Helper functions

```{r}
#| label: functions

season_of <- function(mo) case_when(
  mo %in% c(12, 1, 2) ~ "winter",
  mo %in% 3:5         ~ "spring",
  mo %in% 6:8         ~ "summer",
  TRUE                ~ "fall")

# monthly anomalies as in the deck: subtract the calendar-month mean over the window
get_anom <- function(nms_i, date_beg = as_date("2002-09-01"), date_end = as_date("2025-08-31")) {
  d_wide |>
    filter(nms == nms_i, date >= date_beg, date <= date_end) |>
    select(date, all_of(VARS)) |>
    drop_na() |>
    mutate(mo = month(date)) |>
    group_by(mo) |>
    mutate(across(all_of(VARS), \(x) x - mean(x))) |>
    ungroup() |>
    mutate(season = season_of(mo))
}

# aem basis for a regular series of n points: the first k_max eigenfunctions (the deck uses
# aem 1-24 for the 69-point seasonal subsets) or, if k_max is NULL, all with positive moran's i
aem_cache <- new.env()
get_aem <- function(n, k_max = 24) {
  key <- paste(n, k_max)
  if (is.null(aem_cache[[key]])) {
    a <- aem.time(n, moran = is.null(k_max))
    E <- as.data.frame(a$aem)
    names(E) <- paste0("AEM", seq_len(ncol(E)))
    aem_cache[[key]] <- if (is.null(k_max)) E[, a$Moran$obs > -1 / (n - 1), drop = FALSE] else E[, seq_len(k_max), drop = FALSE]
  }
  aem_cache[[key]]
}

# aem -> forward selection (double stopping criterion) on z-scored responses
fit_aem_rda <- function(Y, nperm = 999, alpha = 0.05, k_max = 24) {
  E   <- get_aem(nrow(Y), k_max)
  Ys  <- scale(Y)
  r2g <- RsquareAdj(rda(Ys, E))$adj.r.squared
  fs  <- tryCatch(
    forward.sel(Ys, E, alpha = alpha, adjR2thresh = r2g, nperm = nperm, verbose = FALSE),
    error = \(e) NULL)
  list(Ys = Ys, E = E, sel = fs, n = nrow(Y))
}

# rda on the leading selected aems with p <= alpha (selection order does not depend on alpha,
# so a stricter alpha is a prefix of the alpha = 0.05 selection)
get_rda <- function(fit, alpha = 0.05) {
  if (is.null(fit$sel)) return(NULL)
  keep <- cumsum(fit$sel$pvalue > alpha) == 0
  if (!any(keep)) return(NULL)
  rda(fit$Ys, fit$E[, fit$sel$variables[keep], drop = FALSE])
}

# indices where two curves cross (the deck's regime boundaries)
crossings <- function(a, b) {
  s <- sign(a - b)
  s <- s[s != 0]
  which(diff(s) != 0)
}

# summary of a fit at each alpha, incl. regimes under both sign conventions for axis ii
summarize_fit <- function(fit, alphas = c(0.05, 0.02)) {
  map_dfr(alphas, \(al) {
    m <- get_rda(fit, al)
    if (is.null(m))
      return(tibble(alpha = al, r2a = 0, n_aem = 0L, aems = "", n_reg = 1L, n_reg_flip = 1L, cross = "", cross_flip = ""))
    k    <- sum(cumsum(fit$sel$pvalue > al) == 0)
    aems <- paste(fit$sel$order[seq_len(k)], collapse = ",")
    r2a  <- RsquareAdj(m)$adj.r.squared
    if (length(eigenvals(m, model = "constrained")) < 2)
      return(tibble(alpha = al, r2a = r2a, n_aem = k, aems = aems, n_reg = 1L, n_reg_flip = 1L, cross = "", cross_flip = ""))
    lc <- scores(m, display = "lc", choices = 1:2, scaling = 1)
    i1 <- crossings(lc[, 1],  lc[, 2])
    i2 <- crossings(lc[, 1], -lc[, 2])
    tibble(
      alpha = al, r2a = r2a, n_aem = k, aems = aems,
      n_reg = length(i1) + 1L, n_reg_flip = length(i2) + 1L,
      cross = paste(i1 + 1, collapse = ","), cross_flip = paste(i2 + 1, collapse = ","))
  })
}

# surrogates with no time-localized shift by construction ----
detrend <- function(X) {
  t   <- seq_len(nrow(X))
  fit <- lm(X ~ t)
  list(trend = fitted(fit), resid = residuals(fit))
}
# linear trend + ar(1) noise with cross-correlated innovations
sur_ar1 <- function(X) {
  dt  <- detrend(X); R <- dt$resid; n <- nrow(R); p <- ncol(R)
  phi <- apply(R, 2, \(x) acf(x, lag.max = 1, plot = FALSE)$acf[2])
  inn <- R[-1, ] - sweep(R[-n, ], 2, phi, `*`)
  Z   <- matrix(rnorm((n + 50) * p), ncol = p) %*% chol(cov(inn))
  S   <- matrix(0, n + 50, p)
  for (i in 2:(n + 50)) S[i, ] <- phi * S[i - 1, ] + Z[i, ]
  dt$trend + S[-(1:50), ]
}
# linear trend + phase-randomized residuals (same random phases for all variables,
# so every auto- and cross-spectrum is preserved but nothing is localized in time)
sur_phase <- function(X) {
  dt  <- detrend(X); R <- dt$resid; n <- nrow(R)
  k   <- floor((n - 1) / 2)
  ph  <- runif(k, 0, 2 * pi)
  rot <- rep(1 + 0i, n)
  rot[2:(k + 1)]     <- exp(1i * ph)
  rot[n:(n - k + 1)] <- exp(-1i * ph)
  dt$trend + Re(mvfft(mvfft(R) * rot, inverse = TRUE)) / n
}

# calendar-month standardized anomalies over the full common record (1998-01..2025-09):
# removes the seasonal cycle in both mean and variance without splitting the series
get_z <- function() {
  d_wide |>
    filter(date >= as_date("1998-01-01"), date <= as_date("2025-09-30")) |>
    select(nms, date, all_of(VARS)) |>
    drop_na() |>
    mutate(
      mo = month(date),
      yr = year(date),
      across(c(CHL, PP, mixThick), log),
      precip = sqrt(precip)) |>
    group_by(nms, mo) |>
    mutate(across(all_of(VARS), \(x) {
      s <- sd(x)
      if (s < 1e-8) rep(0, length(x)) else (x - mean(x)) / s })) |>
    ungroup()
}
z <- get_z()
```

## 1. What the AEM basis is

For a regular series, AEM *k* crosses zero *k* times: it completes one full oscillation in 2*L*/*k*
years, where *L* is the record length. The deck labels AEM *k* as *L*/*k* years, which is the length
of one monotonic excursion (half a cycle). AEM 1 is therefore a monotonic curve that is almost
indistinguishable from a straight line, and AEM 2 completes exactly one oscillation in the record.
The eigenvalues fall as 1/*k*², which is the spectrum of a random walk.

```{r}
#| label: aem_basis

n_s  <- 69                                   # jun-aug 2003-2025
L_yr <- 22.2
a69  <- aem.time(n_s, moran = FALSE)
tibble(
  AEM                 = 1:8,
  `zero crossings`    = apply(a69$aem[, 1:8], 2, \(v) sum(diff(sign(v)) != 0)),
  `deck label (yr)`   = round(L_yr / (1:8), 1),
  `full cycle (yr)`   = round(2 * L_yr / (1:8), 1),
  `eigenvalue / first` = round(a69$values[1:8] / a69$values[1], 3),
  `1 / k^2`           = round(1 / (1:8)^2, 3)) |>
  kable(caption = glue(
    "AEM basis for n = {n_s}. Correlation of AEM 1 with time = ",
    "{round(abs(cor(a69$aem[, 1], seq_len(n_s))), 3)}."))
```

```{r}
#| label: fig_aem_basis
#| fig-width: 9
#| fig-height: 3.2

d_aem <- as_tibble(a69$aem[, 1:4]) |>
  setNames(glue("AEM {1:4}\ndeck label {round(L_yr / 1:4, 1)} yr; full cycle {round(2 * L_yr / 1:4, 1)} yr")) |>
  mutate(year = 2003 + (row_number() - 1) / 3) |>
  pivot_longer(-year, names_to = "aem")

p <- ggplot(d_aem, aes(year, value)) +
  geom_hline(yintercept = 0, color = PAL$axis, linewidth = 0.3) +
  geom_line(color = PAL$blue, linewidth = 0.7) +
  facet_wrap(~aem, nrow = 1) +
  labs(
    title = "AEM 1 is a trend, not a cycle",
    subtitle = "First four temporal eigenfunctions for a 69-point (22.2 yr) seasonal subset",
    x = NULL, y = "eigenfunction value") +
  theme_chk()
save_fig(p, "fig_aem_basis.png", 9, 3.2)
```

## 2. Replication of the flagship analysis (HIHWNMS, summer)

Candidates are AEM 1–24 as in the deck. Forward selection at α = 0.05 retains slightly more
eigenfunctions than the deck; α = 0.02 reproduces the deck's 52 models most closely (section 4), so
both are carried through.

```{r}
#| label: replicate

ALPHA_DECK <- 0.02      # reproduces the deck's models most closely

a_hi   <- get_anom("HIHWNMS") |> filter(season == "summer")
fit_hi <- fit_aem_rda(as.matrix(a_hi[, VARS]))
m_hi   <- get_rda(fit_hi, ALPHA_DECK)

deck_hi <- tibble(
  variables  = paste0("AEM", c(1, 3, 2, 14, 4, 8, 16, 5)),
  deck_R2adj = c(0.1527, 0.1399, 0.0554, 0.0309, 0.0299, 0.0169, 0.0154, 0.0118))

fit_hi$sel |>
  as_tibble() |>
  mutate(this_R2adj = c(AdjR2Cum[1], diff(AdjR2Cum))) |>
  select(variables, this_R2adj, pvalue) |>
  full_join(deck_hi, by = "variables") |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  kable(caption = glue(
    "Forward-selected AEMs (alpha = 0.05), this notebook vs. the deck (slide 12). n = {nrow(a_hi)}. ",
    "Total adjusted R2: {round(RsquareAdj(get_rda(fit_hi, 0.05))$adj.r.squared, 3)} at alpha 0.05, ",
    "{round(RsquareAdj(m_hi)$adj.r.squared, 3)} at alpha {ALPHA_DECK}; deck 0.453. ",
    "Differences reflect the refreshed extract and unspecified pre-processing details."))
```

## 3. Are axis-crossing regimes invariant?

The control chart draws the fitted site scores of canonical axes I and II through time and places a
regime boundary wherever the two curves cross. Eigenvectors have no intrinsic sign: reflecting an
axis gives an equally valid ordination. If boundaries move when an axis is reflected, they describe a
plotting convention rather than the system.

```{r}
#| label: crossing_conventions

get_regimes <- function(m, dates, scaling = 1, flip2 = 1) {
  lc <- scores(m, display = "lc", choices = 1:2, scaling = scaling)
  i  <- crossings(lc[, 1], flip2 * lc[, 2])
  tibble(
    scaling    = scaling,
    `axis II`  = if_else(flip2 == 1, "as returned", "reflected"),
    regimes    = length(i) + 1,
    boundaries = paste(format(dates[i + 1], "%Y-%m"), collapse = ", "))
}
expand_grid(alpha = c(ALPHA_DECK, 0.05), scaling = c(1, 2), flip2 = c(1, -1)) |>
  pmap_dfr(\(alpha, scaling, flip2)
    get_regimes(get_rda(fit_hi, alpha), a_hi$date, scaling, flip2) |> mutate(alpha = alpha, .before = 1)) |>
  kable(caption = "HIHWNMS summer: regime boundaries under equivalent presentations of the same RDA. The deck reports boundaries at Jul 2009, Jul 2010 and Jul 2018.")
```

```{r}
#| label: fig_control_chart_sign
#| fig-width: 9
#| fig-height: 5.6

lc <- scores(m_hi, display = "lc", choices = 1:2, scaling = 1)
conv_lab <- c(
  a = "Convention A: axis II as returned by the software",
  b = "Convention B: axis II reflected (an equally valid solution)")
d_cc <- bind_rows(
  tibble(conv = conv_lab["a"], date = a_hi$date, `Axis I` = lc[, 1], `Axis II` =  lc[, 2]),
  tibble(conv = conv_lab["b"], date = a_hi$date, `Axis I` = lc[, 1], `Axis II` = -lc[, 2])) |>
  pivot_longer(c(`Axis I`, `Axis II`), names_to = "axis", values_to = "score")
d_bd <- bind_rows(
  tibble(conv = conv_lab["a"], date = a_hi$date[crossings(lc[, 1],  lc[, 2]) + 1]),
  tibble(conv = conv_lab["b"], date = a_hi$date[crossings(lc[, 1], -lc[, 2]) + 1])) |>
  mutate(lab = format(date, "%Y"))
y_top <- max(abs(d_cc$score)) * 1.25

p <- ggplot(d_cc, aes(date, score)) +
  geom_vline(data = d_bd, aes(xintercept = date), color = PAL$ink2, linewidth = 0.3) +
  geom_text(data = d_bd, aes(x = date, y = y_top, label = lab),
            color = PAL$ink2, size = 2.8, hjust = -0.15, check_overlap = TRUE) +
  geom_line(aes(color = axis), linewidth = 0.7) +
  facet_wrap(~conv, ncol = 1) +
  scale_color_manual(values = c("Axis I" = PAL$blue, "Axis II" = PAL$orange), name = NULL) +
  labs(
    title = "Regime boundaries move when an axis is reflected",
    subtitle = "HIHWNMS summer anomalies: fitted site scores of canonical axes I and II; vertical lines mark crossings",
    x = NULL, y = "fitted site score (scaling 1)") +
  theme_chk()
save_fig(p, "fig_control_chart_sign.png", 9, 5.6)
```

## 4. Do crossings exceed what "trend + red noise" produces?

All 52 season × sanctuary analyses were re-fit, then compared with surrogate series that contain a
linear trend plus autocorrelated noise and, by construction, **no** time-localized shift:
`trend_ar1` (AR(1) noise with cross-correlated innovations) and `trend_phase` (phase-randomized
residuals that preserve every auto- and cross-spectrum).

```{r}
#| label: null_52

if (!file.exists(null_csv)) {
  jobs <- expand_grid(nms = NMS_ORD, season = SEASONS)
  run_job <- function(j) {
    set.seed(1000 + j)
    a   <- get_anom(jobs$nms[j])
    X   <- as.matrix(a[, VARS])
    idx <- which(a$season == jobs$season[j])
    to_dates <- \(s) map_chr(s, \(x) if (x == "") "" else
      paste(format(a$date[idx][as.integer(str_split(x, ",")[[1]])], "%Y-%m"), collapse = ","))
    fit <- \(M, nperm) summarize_fit(fit_aem_rda(M[idx, ], nperm = nperm))
    bind_rows(
      fit(X, 999) |> mutate(type = "observed", rep = 0L),
      map_dfr(seq_len(N_SUR), \(r) fit(sur_ar1(X),   199) |> mutate(type = "trend_ar1",   rep = r)),
      map_dfr(seq_len(N_SUR), \(r) fit(sur_phase(X), 199) |> mutate(type = "trend_phase", rep = r))) |>
      mutate(
        nms = jobs$nms[j], season = jobs$season[j], n = length(idx),
        cross_dates      = if_else(type == "observed", to_dates(cross), ""),
        cross_dates_flip = if_else(type == "observed", to_dates(cross_flip), ""))
  }
  invisible(get_aem(69))
  mclapply(seq_len(nrow(jobs)), run_job, mc.cores = N_CORE) |>
    bind_rows() |>
    select(nms, season, n, type, rep, alpha, r2a, n_aem, aems, n_reg, n_reg_flip, cross_dates, cross_dates_flip) |>
    write_csv(null_csv)
}
d_null <- read_csv(null_csv, col_types = cols(
  aems = "c", cross_dates = "c", cross_dates_flip = "c")) |>
  mutate(across(c(aems, cross_dates, cross_dates_flip), \(x) replace_na(x, "")))
d_obs  <- d_null |> filter(type == "observed")
d_deck <- read_csv(deck_csv)

# replication fidelity across all 52 analyses, by alpha
d_obs |>
  left_join(d_deck |> rename(n_aem_deck = n_aem), by = c("nms", "season")) |>
  group_by(alpha) |>
  summarize(
    `r (adj. R2, here vs deck)`         = round(cor(r2a, r2adj_pct), 2),
    `median adj. R2 here (%)`           = round(100 * median(r2a), 1),
    `median adj. R2 deck (%)`           = median(r2adj_pct),
    `median abs. difference (points)`   = round(median(abs(100 * r2a - r2adj_pct)), 1),
    `median AEMs here`                  = median(n_aem),
    `median AEMs deck`                  = median(n_aem_deck),
    .groups = "drop") |>
  kable(caption = "How closely this notebook reproduces the deck's 52 models (slide 33)")
```

```{r}
#| label: null_summary

has_aem1 <- \(x) str_detect(paste0(",", x, ","), ",1,")
d_sum <- d_null |>
  filter(type != "observed") |>
  group_by(alpha, nms, season, type) |>
  summarize(
    reg_med = median(n_reg), reg_lo = quantile(n_reg, 0.05), reg_hi = quantile(n_reg, 0.95),
    r2a_med = median(r2a),   r2a_lo = quantile(r2a, 0.05),   r2a_hi = quantile(r2a, 0.95),
    aem1    = mean(has_aem1(aems)),
    .groups = "drop") |>
  left_join(
    d_obs |> select(alpha, nms, season, reg_obs = n_reg, reg_obs_flip = n_reg_flip, r2a_obs = r2a, aems_obs = aems),
    by = c("alpha", "nms", "season"))

d_sum |>
  group_by(alpha, type) |>
  summarize(
    `observed regimes per analysis, median (range)` = glue("{median(reg_obs)} ({min(reg_obs)}-{max(reg_obs)})"),
    `surrogate regimes per analysis, median`        = median(reg_med),
    `observed regime count inside surrogate 5-95%`  = glue("{sum(reg_obs >= reg_lo & reg_obs <= reg_hi)} of {n()}"),
    `observed regime count above surrogate 95%`     = glue("{sum(reg_obs > reg_hi)} of {n()}"),
    `observed adj. R2 above surrogate 95%`          = glue("{sum(r2a_obs > r2a_hi)} of {n()}"),
    `observed analyses selecting AEM 1`             = glue("{sum(has_aem1(aems_obs))} of {n()}"),
    `surrogates selecting AEM 1 (%)`                = round(100 * mean(aem1)),
    .groups = "drop") |>
  kable(caption = glue("Observed analyses vs. no-shift surrogates ({N_SUR} per analysis)"))

d_obs |>
  group_by(alpha) |>
  summarize(
    `regimes, axis II as returned` = glue("{sum(n_reg)} in total; median {median(n_reg)}"),
    `regimes, axis II reflected`   = glue("{sum(n_reg_flip)} in total; median {median(n_reg_flip)}"),
    `analyses with identical boundaries under both conventions` = glue("{sum(cross_dates == cross_dates_flip)} of {n()}"),
    .groups = "drop") |>
  kable(caption = "Sensitivity of crossing-defined regimes to the sign of axis II, all 52 analyses")
```

```{r}
#| label: fig_null_regimes
#| fig-width: 9
#| fig-height: 4.8

d_plot <- d_sum |> filter(type == "trend_phase", alpha == ALPHA_DECK)
lim_r  <- c(0, max(d_plot$reg_hi, d_plot$reg_obs) + 1)
p1 <- ggplot(d_plot, aes(reg_obs, reg_med)) +
  geom_abline(slope = 1, intercept = 0, color = PAL$axis, linewidth = 0.3) +
  geom_linerange(aes(ymin = reg_lo, ymax = reg_hi), color = PAL$axis, linewidth = 0.4,
                 position = position_jitter(width = 0.2, height = 0, seed = 1)) +
  geom_point(color = PAL$blue, size = 1.8,
             position = position_jitter(width = 0.2, height = 0, seed = 1)) +
  coord_equal(xlim = lim_r, ylim = lim_r) +
  labs(
    title = "Number of crossing-defined regimes",
    x = "observed data", y = "no-shift surrogates (median, 5-95%)") +
  theme_chk()
lim_a <- c(0, max(d_plot$r2a_hi, d_plot$r2a_obs) + 0.05)
p2 <- ggplot(d_plot, aes(r2a_obs, r2a_med)) +
  geom_abline(slope = 1, intercept = 0, color = PAL$axis, linewidth = 0.3) +
  geom_linerange(aes(ymin = r2a_lo, ymax = r2a_hi), color = PAL$axis, linewidth = 0.4) +
  geom_point(color = PAL$blue, size = 1.8) +
  scale_x_continuous(labels = label_percent()) +
  scale_y_continuous(labels = label_percent()) +
  coord_equal(xlim = lim_a, ylim = lim_a) +
  labs(
    title = "Adjusted R² of the AEM model",
    x = "observed data", y = "no-shift surrogates (median, 5-95%)") +
  theme_chk()
p <- (p1 | p2) +
  plot_annotation(
    title = "Series with no shift produce as many axis-crossing \"regimes\" and as much temporal structure",
    subtitle = glue(
      "Each point is one of 52 season × sanctuary analyses (alpha = {ALPHA_DECK}); grey line = 1:1;\n",
      "surrogates = linear trend + phase-randomized residuals, {N_SUR} per analysis"),
    theme = theme_chk())
save_fig(p, "fig_null_regimes.png", 9, 4.8)
```

## 5. Do known events appear in sanctuary-mean monthly anomalies?

Anomalies here are standardized by calendar month over 1998-01 to 2025-09 (chlorophyll, primary
production and mixed-layer depth log-transformed; precipitation square-root-transformed), so the
seasonal cycle is removed in both mean and variance without splitting the series. Event windows are
from the manuscript's event catalog; they are **not** used in computing anything shown.

```{r}
#| label: events

# lab_row staggers labels of events that are close in time
events <- tribble(
  ~nms,      ~event,                                   ~beg,         ~end,         ~lab_row,
  "CINMS",   "1997-98 El Niño",                        "1998-01-01", "1998-06-30", 1,
  "CINMS",   "2014-16 marine heatwave",                "2014-01-01", "2016-08-31", 1,
  "HIHWNMS", "2015 heatwave",                          "2015-07-01", "2015-11-30", 1,
  "HIHWNMS", "2019 heatwave",                          "2019-06-01", "2019-12-31", 1,
  "FKNMS",   "2010 cold winter (Jan cold snap)",       "2010-01-01", "2010-03-31", 1,
  "FKNMS",   "2023 heatwave",                          "2023-06-01", "2023-09-30", 1,
  "FGBNMS",  "Hurricane Ike (Sep 2008)",               "2008-09-01", "2008-10-31", 2,
  "FGBNMS",  "2010 cold winter",                       "2010-01-01", "2010-04-30", 1,
  "FGBNMS",  "Jul 2016 mortality event (hypoxia)",     "2016-07-01", "2016-07-31", 2,
  "FGBNMS",  "Hurricane Harvey (Sep 2017)",            "2017-09-01", "2017-10-31", 1,
  "GRNMS",   "2003 summer cold event",                 "2003-07-01", "2003-08-31", 1,
  "GRNMS",   "2010 cold winter",                       "2010-01-01", "2010-03-31", 1,
  "SBNMS",   "2012 heatwave",                          "2012-01-01", "2012-12-31", 1) |>
  mutate(across(c(beg, end), as_date))

sigma_from_d2 <- function(d2, p) qnorm(pchisq(d2, df = p) / 2 + 0.5)
z <- z |>
  group_by(nms) |>
  group_modify(\(d, k) {
    X  <- as.matrix(d[, VARS])
    d2 <- mahalanobis(X, colMeans(X), cov(X))
    d |> mutate(sigma = sigma_from_d2(d2, length(VARS)), rank_mv = rank(-d2))
  }) |>
  ungroup()

events |>
  rowwise() |>
  mutate(s = list(
    z |>
      filter(nms == .env$nms, date >= beg, date <= end) |>
      summarize(
        months            = n(),
        across(all_of(VARS), \(x) x[which.max(abs(x))]),
        `best rank of 333` = min(rank_mv)))) |>
  unnest(s) |>
  mutate(across(all_of(VARS), \(x) round(x, 1)), window = glue("{format(beg, '%Y-%m')} to {format(end, '%Y-%m')}")) |>
  select(nms, event, window, months, all_of(VARS), `best rank of 333`) |>
  kable(caption = "Most extreme standardized anomaly (SD units) of each variable inside each event window, and the best multivariate rank (1 = most unusual month of 333)")
```

```{r}
#| label: fig_event_strips
#| fig-width: 10
#| fig-height: 10

nms_ev <- c("CINMS", "HIHWNMS", "FKNMS", "FGBNMS", "GRNMS", "SBNMS")
d_strip <- z |>
  filter(nms %in% nms_ev) |>
  select(nms, date, all_of(VARS)) |>
  pivot_longer(all_of(VARS), names_to = "var", values_to = "z") |>
  mutate(
    nms = factor(nms, nms_ev),
    var = factor(var, rev(VARS)),
    x   = year(date) + (month(date) - 0.5) / 12,
    z   = pmax(pmin(z, 3), -3))
d_ev <- events |>
  mutate(
    nms  = factor(nms, nms_ev),
    xmin = year(beg) + (month(beg) - 1) / 12,
    xmax = year(end) + month(end) / 12)

p <- ggplot(d_strip, aes(x, as.numeric(var))) +
  geom_raster(aes(fill = z)) +
  geom_rect(data = d_ev, inherit.aes = FALSE,
            aes(xmin = xmin, xmax = xmax, ymin = 0.5, ymax = 7.5),
            fill = NA, color = PAL$ink, linewidth = 0.35) +
  geom_text(data = d_ev, inherit.aes = FALSE,
            aes(x = xmin, y = 7.35 + 0.8 * lab_row, label = event),
            hjust = 0, size = 2.7, color = PAL$ink2) +
  facet_wrap(~nms, ncol = 1) +
  scale_fill_gradientn(
    colors = PAL_DIV, limits = c(-3, 3), breaks = -3:3,
    name = "standardized anomaly (SD, capped at ±3)",
    guide = guide_colorbar(barwidth = 14, barheight = 0.5, title.position = "top")) +
  scale_y_continuous(breaks = seq_along(VARS), labels = rev(VARS), limits = c(0.5, 9.4), expand = c(0, 0)) +
  scale_x_continuous(breaks = seq(1998, 2026, 2), expand = c(0, 0)) +
  labs(
    title = "Large, persistent events are unmistakable in sanctuary-mean monthly anomalies; short or local ones are not",
    subtitle = "Calendar-month standardized anomalies, 1998-2025; boxes mark independently documented event windows",
    x = NULL, y = NULL) +
  theme_chk() +
  theme(panel.grid = element_blank())
save_fig(p, "fig_event_strips.png", 10, 10)
```

## 6. One analysis per sanctuary instead of four

Running the same AEM–RDA on the full monthly series (n = 333) resolves sub-annual to multi-decadal
structure in one model per sanctuary. Selected eigenfunctions are grouped into four time-scale bands
by their full-cycle period (2*L*/*k*).

```{r}
#| label: full_series

if (!file.exists(full_csv)) {
  fit_full <- function(nms_i) {
    d   <- z |> filter(nms == nms_i) |> arrange(date)
    Y   <- as.matrix(d[, VARS])
    E   <- get_aem(nrow(Y), k_max = NULL)
    r2g <- RsquareAdj(rda(Y, E))$adj.r.squared
    fs  <- forward.sel(Y, E, alpha = 0.05, adjR2thresh = r2g, nperm = 499, verbose = FALSE)
    tibble(nms = nms_i, n = nrow(Y), k = fs$order, r2a_inc = c(fs$AdjR2Cum[1], diff(fs$AdjR2Cum)))
  }
  invisible(get_aem(333, k_max = NULL))
  mclapply(NMS_ORD, fit_full, mc.cores = N_CORE) |>
    bind_rows() |>
    write_csv(full_csv)
}
BANDS <- c("< 1.5 yr", "1.5-8 yr (interannual)", "8-15 yr (decadal)", "> 15 yr (trend-like)")
d_full <- read_csv(full_csv) |>
  mutate(
    period_yr = 2 * (n / 12) / k,
    band      = cut(period_yr, c(0, 1.5, 8, 15, Inf), labels = BANDS),
    nms       = factor(nms, rev(NMS_ORD)))
d_band <- d_full |>
  group_by(nms, band) |>
  summarize(r2a = sum(r2a_inc), .groups = "drop")
d_tot <- d_full |>
  group_by(nms) |>
  summarize(r2a = sum(r2a_inc), n_aem = n(), .groups = "drop")

d_band |>
  mutate(r2a = round(100 * r2a, 1)) |>
  pivot_wider(names_from = band, values_from = r2a, values_fill = 0, names_sort = T) |>
  left_join(d_tot |> mutate(`total adj. R2 (%)` = round(100 * r2a, 1)) |> select(nms, `total adj. R2 (%)`, `AEMs selected` = n_aem), by = "nms") |>
  arrange(desc(nms)) |>
  kable(caption = "Full-series AEM–RDA (n = 333 months per sanctuary): adjusted R² (%) by time-scale band")
```

```{r}
#| label: fig_timescale_bands
#| fig-width: 9
#| fig-height: 4.8

p <- ggplot(d_band, aes(r2a, nms)) +
  geom_col(aes(fill = band), width = 0.62, color = PAL$surface, linewidth = 0.6,
           position = position_stack(reverse = TRUE)) +
  geom_text(data = d_tot, aes(x = r2a, label = percent(r2a, accuracy = 1)),
            hjust = -0.25, size = 3, color = PAL$ink2) +
  scale_fill_manual(values = setNames(PAL_BAND, BANDS), name = "time-scale band (full-cycle period)") +
  scale_x_continuous(labels = label_percent(), expand = expansion(mult = c(0, 0.08))) +
  labs(
    title = "Time-scale fingerprint of each sanctuary from one full-series model",
    subtitle = "Adjusted R² of the 7-variable anomaly state explained by temporal eigenfunctions, 1998-2025 (n = 333 months)",
    x = "adjusted R²", y = NULL) +
  theme_chk() +
  theme(panel.grid.major.y = element_blank())
save_fig(p, "fig_timescale_bands.png", 9, 4.8)
```

### Distribution-free change points and a chronology prototype

E-divisive (`ecp::e.divisive`) finds multivariate change points without reference to ordination
axes. Annual means are used for the chronology because monthly anomalies are autocorrelated, which
inflates detections.

```{r}
#| label: changepoints

if (!file.exists(cp_csv)) {
  cp_monthly <- function(nms_i) {
    d  <- z |> filter(nms == nms_i) |> arrange(date)
    i  <- e.divisive(as.matrix(d[, VARS]), sig.lvl = 0.01, R = 499, min.size = 36)$estimates
    i  <- i[i > 1 & i <= nrow(d)]
    tibble(nms = nms_i, cp_monthly = paste(format(d$date[i], "%Y-%m"), collapse = ", "))
  }
  cp_annual <- function(nms_i) {
    d  <- z |> filter(nms == nms_i, yr <= 2024) |>
      group_by(yr) |> summarize(across(all_of(VARS), mean), .groups = "drop")
    i  <- e.divisive(as.matrix(d[, VARS]), sig.lvl = 0.05, R = 999, min.size = 4)$estimates
    i  <- i[i > 1 & i <= nrow(d)]
    tibble(nms = nms_i, cp_annual = paste(d$yr[i], collapse = ", "))
  }
  left_join(
    mclapply(NMS_ORD, cp_monthly, mc.cores = N_CORE) |> bind_rows(),
    mclapply(NMS_ORD, cp_annual,  mc.cores = N_CORE) |> bind_rows(), by = "nms") |>
    write_csv(cp_csv)
}
d_cp <- read_csv(cp_csv, col_types = "ccc")
d_cp |>
  mutate(nms = factor(nms, NMS_ORD)) |>
  arrange(nms) |>
  kable(caption = "E-divisive change points: monthly anomalies (min. segment 36 months, alpha 0.01) and annual means (min. 4 years, alpha 0.05; value = first year of the new segment)")
```

```{r}
#| label: fig_chronology
#| fig-width: 9.5
#| fig-height: 4.8

ann <- z |>
  group_by(nms, yr) |>
  summarize(across(all_of(VARS), mean), .groups = "drop") |>
  group_by(nms) |>
  mutate(across(all_of(VARS), \(x) x / sd(x))) |>
  ungroup()
d_chr <- ann |>
  mutate(
    rms = sqrt(rowMeans(across(all_of(VARS), \(x) x^2))),
    nms = factor(nms, rev(NMS_ORD)))
d_cpa <- d_cp |>
  separate_longer_delim(cp_annual, ", ") |>
  filter(!is.na(cp_annual), cp_annual != "") |>
  mutate(yr = as.numeric(cp_annual), nms = factor(nms, rev(NMS_ORD)))

p <- ggplot(d_chr, aes(yr, nms)) +
  geom_tile(aes(fill = rms), color = PAL$surface, linewidth = 0.6) +
  geom_segment(data = d_cpa, inherit.aes = FALSE,
               aes(x = yr - 0.5, xend = yr - 0.5, y = as.numeric(nms) - 0.48, yend = as.numeric(nms) + 0.48),
               color = PAL$ink, linewidth = 0.9) +
  scale_fill_gradientn(
    colors = PAL_SEQ, limits = c(0, NA),
    name = "departure of the annual state (RMS of 7 standardized annual anomalies)",
    guide = guide_colorbar(barwidth = 14, barheight = 0.5, title.position = "top")) +
  scale_x_continuous(breaks = seq(1998, 2025, 3), expand = c(0, 0)) +
  labs(
    title = "Chronology prototype: when each sanctuary's environmental state departed, and where it changed",
    subtitle = "Tiles = size of the annual multivariate anomaly; black bars = e-divisive change points on annual means",
    x = NULL, y = NULL) +
  theme_chk() +
  theme(panel.grid = element_blank())
save_fig(p, "fig_chronology.png", 9.5, 4.8)
```

### How far has each sanctuary moved from its own baseline?

Shift of the most recent decade (2016-2025) from a baseline decade, per variable, in units of the
baseline's interannual standard deviation. Units are dimensionless, so sanctuaries can be compared.

```{r}
#| label: departure

get_departure <- function(yr_base) {
  z |>
    group_by(nms, yr) |>
    summarize(across(all_of(VARS), mean), .groups = "drop") |>
    group_by(nms) |>
    group_modify(\(d, k) {
      B  <- as.matrix(d[d$yr %in% yr_base, VARS])
      R  <- as.matrix(d[d$yr >= 2016, VARS])
      sh <- (colMeans(R) - colMeans(B)) / apply(B, 2, sd)
      as_tibble(as.list(sh)) |> mutate(rms = sqrt(mean(sh^2)))
    }) |>
    ungroup() |>
    mutate(baseline = glue("{min(yr_base)}-{max(yr_base)}"))
}
d_dep <- bind_rows(get_departure(1998:2007), get_departure(2003:2012))

d_dep |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(baseline, desc(rms)) |>
  select(baseline, nms, all_of(VARS), `RMS shift` = rms) |>
  kable(caption = "Shift of the 2016-2025 mean from two alternative baseline decades (baseline interannual SD units)")
```

```{r}
#| label: fig_departure
#| fig-width: 9.5
#| fig-height: 4.8

d1    <- d_dep |> filter(baseline == "2003-2012")
ord   <- d1 |> arrange(rms) |> pull(nms)
d_var <- d1 |>
  pivot_longer(all_of(VARS), names_to = "var", values_to = "shift") |>
  mutate(nms = factor(nms, ord), var = factor(var, VARS))

p1 <- ggplot(d_var, aes(var, nms)) +
  geom_tile(aes(fill = pmax(pmin(shift, 4), -4)), color = PAL$surface, linewidth = 0.8) +
  geom_text(aes(label = sprintf("%+.1f", shift),
                color = abs(shift) > 2.4), size = 2.9, show.legend = FALSE) +
  scale_color_manual(values = c(`TRUE` = "white", `FALSE` = PAL$ink)) +
  scale_fill_gradientn(
    colors = PAL_DIV, limits = c(-4, 4), name = "shift (baseline SD)",
    guide = guide_colorbar(barwidth = 9, barheight = 0.5, title.position = "top")) +
  scale_x_discrete(position = "top") +
  labs(x = NULL, y = NULL) +
  theme_chk() +
  theme(panel.grid = element_blank())
p2 <- ggplot(d1 |> mutate(nms = factor(nms, ord)), aes(rms, nms)) +
  geom_col(fill = PAL$blue, width = 0.5) +
  geom_text(aes(label = sprintf("%.1f", rms)), hjust = -0.3, size = 2.9, color = PAL$ink2) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = "RMS shift across 7 variables", y = NULL) +
  theme_chk() +
  theme(axis.text.y = element_blank(), panel.grid.major.y = element_blank())
p <- (p1 | p2) +
  plot_layout(widths = c(3, 1.2)) +
  plot_annotation(
    title = "Which sanctuary changed most relative to itself, and which variables drove it",
    subtitle = "Mean of 2016-2025 minus mean of 2003-2012, in units of each sanctuary's own 2003-2012 interannual SD",
    theme = theme_chk())
save_fig(p, "fig_departure.png", 9.5, 4.8)
```

## 7. A shared step in temperature

Surface temperature is available from 1993 and does not depend on the ocean-color record. For each
sanctuary, a single-step model of the annual anomaly (break year estimated, with a BIC penalty for
estimating it) is compared with a linear trend.

```{r}
#| label: sst_step

d_sst <- d_wide |>
  filter(date >= as_date("1993-01-01"), date <= as_date("2025-12-31")) |>
  select(nms, date, tempSurf) |>
  drop_na() |>
  mutate(mo = month(date), yr = year(date)) |>
  group_by(nms, mo) |>
  mutate(anom = tempSurf - mean(tempSurf[yr <= 2012])) |>
  group_by(nms, yr) |>
  summarize(anom = mean(anom), n = n(), .groups = "drop") |>
  filter(n == 12)

d_step <- d_sst |>
  group_by(nms) |>
  group_modify(\(d, k) {
    n     <- nrow(d)
    cand  <- d$yr[6:(n - 4)]                       # at least 5 years on either side
    bic_s <- sapply(cand, \(y) BIC(lm(anom ~ I(yr >= y), d)) + log(n))
    y     <- cand[which.min(bic_s)]
    m_tr  <- lm(anom ~ yr, d)
    pre   <- d$anom[d$yr < y]
    post  <- d$anom[d$yr >= y]
    tibble(
      step_year  = y,
      step_C     = mean(post) - mean(pre),
      step_sd    = (mean(post) - mean(pre)) / sd(pre),
      mean_pre   = mean(pre),
      mean_post  = mean(post),
      yrs_above_pre_max = glue("{sum(post > max(pre))} of {length(post)}"),
      bic_trend  = BIC(m_tr),
      bic_step   = min(bic_s),
      better     = case_when(
        bic_step < bic_trend - 2 ~ "step",
        bic_trend < bic_step - 2 ~ "trend",
        TRUE                     ~ "neither"))
  }) |>
  ungroup() |>
  mutate(nms = factor(nms, NMS_ORD)) |>
  arrange(nms)

d_step |>
  mutate(across(c(step_C, step_sd, bic_trend, bic_step), \(x) round(x, 2))) |>
  select(nms, `best step year` = step_year, `step (°C)` = step_C, `step (pre-step SD)` = step_sd,
         `post-step years above pre-step maximum` = yrs_above_pre_max,
         `BIC trend` = bic_trend, `BIC step` = bic_step, `better model` = better) |>
  kable(caption = "Annual surface-temperature anomaly, 1993-2025 (vs 1993-2012 monthly climatology): single step vs linear trend")
```

Is a shared step distinguishable from a linear warming trend plus red noise? Surrogates keep each
sanctuary's linear trend, the spectrum of its residuals and the correlations among sanctuaries
(common random phases), and contain no step.

```{r}
#| label: sst_step_null

step_null_csv <- file.path(dir_chk, "sst_step_null.csv")
X_sst <- d_sst |>
  select(nms, yr, anom) |>
  pivot_wider(names_from = nms, values_from = anom) |>
  arrange(yr)
yrs   <- X_sst$yr
X_sst <- as.matrix(X_sst[, NMS_ORD])
n_y   <- nrow(X_sst)
cand  <- 6:(n_y - 4)

# best single-step year and delta bic (trend minus step; positive favors a step)
step_stat <- function(x) {
  bic    <- \(rss, k) n_y * log(rss / n_y) + k * log(n_y)
  rss_tr <- sum(resid(lm(x ~ yrs))^2)
  rss_st <- sapply(cand, \(i) sum((x[1:(i - 1)] - mean(x[1:(i - 1)]))^2) + sum((x[i:n_y] - mean(x[i:n_y]))^2))
  j      <- which.min(rss_st)
  c(year = yrs[cand[j]], dbic = bic(rss_tr, 3) - bic(rss_st[j], 4))
}
obs_step <- apply(X_sst, 2, step_stat)

if (!file.exists(step_null_csv)) {
  set.seed(42)
  fit_tr <- lm(X_sst ~ yrs)
  Tr     <- fitted(fit_tr)
  R      <- resid(fit_tr)
  sur    <- function() {
    k   <- floor((n_y - 1) / 2)
    ph  <- runif(k, 0, 2 * pi)
    rot <- rep(1 + 0i, n_y)
    rot[2:(k + 1)]       <- exp(1i * ph)
    rot[n_y:(n_y - k + 1)] <- exp(-1i * ph)
    Tr + Re(mvfft(mvfft(R) * rot, inverse = TRUE)) / n_y
  }
  map_dfr(seq_len(5000), \(b) {
    S <- apply(sur(), 2, step_stat)
    as_tibble(as.list(S["dbic", ])) |>
      mutate(
        rep     = b,
        modal   = max(table(S["year", ])),
        n_step  = sum(S["dbic", ] > 2),
        n_2014  = sum(S["year", ] == 2014))
  }) |>
    write_csv(step_null_csv)
}
d_sn <- read_csv(step_null_csv)

tibble(
  nms          = NMS_ORD,
  `step year`  = obs_step["year", ],
  `delta BIC`  = round(obs_step["dbic", ], 1),
  `surrogate p` = map_dbl(NMS_ORD, \(x) mean(d_sn[[x]] >= obs_step["dbic", x]))) |>
  mutate(`surrogate p` = round(`surrogate p`, 3)) |>
  kable(caption = "Step vs trend per sanctuary: share of trend-plus-red-noise surrogates with a delta BIC at least as large as observed")

obs_modal <- max(table(obs_step["year", ]))
obs_nstep <- sum(obs_step["dbic", ] > 2)
tibble(
  statistic = c(
    "sanctuaries sharing the modal step year",
    "sanctuaries where BIC prefers a step (delta BIC > 2)",
    "sum of delta BIC over sanctuaries"),
  observed  = c(obs_modal, obs_nstep, round(sum(obs_step["dbic", ]), 1)),
  `surrogate median` = c(median(d_sn$modal), median(d_sn$n_step), round(median(rowSums(d_sn[, NMS_ORD])), 1)),
  `surrogate 95th percentile` = c(
    quantile(d_sn$modal, 0.95), quantile(d_sn$n_step, 0.95), round(quantile(rowSums(d_sn[, NMS_ORD]), 0.95), 1)),
  `surrogate p` = c(
    mean(d_sn$modal >= obs_modal), mean(d_sn$n_step >= obs_nstep),
    mean(rowSums(d_sn[, NMS_ORD]) >= sum(obs_step["dbic", ])))) |>
  kable(caption = "System-level test of a shared step (5000 surrogates)")
```

```{r}
#| label: fig_sst_step
#| fig-width: 10
#| fig-height: 6.4

d_seg <- d_step |>
  transmute(
    nms, step_year,
    lab = glue("{step_year}: {sprintf('%+.2f', step_C)} °C ({better})"),
    x1 = 1992.5, x2 = step_year - 0.5, x3 = 2025.5, mean_pre, mean_post)

p <- d_sst |>
  mutate(nms = factor(nms, NMS_ORD)) |>
  ggplot(aes(yr, anom)) +
  geom_hline(yintercept = 0, color = PAL$axis, linewidth = 0.3) +
  geom_line(color = PAL$blue, linewidth = 0.6) +
  geom_point(color = PAL$blue, size = 0.9) +
  geom_segment(data = d_seg, aes(x = x1, xend = x2, y = mean_pre,  yend = mean_pre),  color = PAL$ink, linewidth = 0.6) +
  geom_segment(data = d_seg, aes(x = x2, xend = x3, y = mean_post, yend = mean_post), color = PAL$ink, linewidth = 0.6) +
  geom_text(data = d_seg, aes(x = 1993, y = Inf, label = lab), hjust = 0, vjust = 1.4, size = 2.8, color = PAL$ink2) +
  facet_wrap(~nms, ncol = 5) +
  scale_x_continuous(breaks = c(1995, 2005, 2015, 2025)) +
  labs(
    title = "A step rather than a ramp: surface temperature shifted abruptly near 2014 in most sanctuaries",
    subtitle = "Annual anomaly (°C) from the 1993-2012 monthly climatology; black lines = means before and after the best-fitting step year;\nlabel = step year, step size and which model BIC prefers (step vs linear trend)",
    x = NULL, y = "annual surface-temperature anomaly (°C)") +
  theme_chk()
save_fig(p, "fig_sst_step.png", 10, 6.4)
```

## 8. Product checks that bear on "regime shifts"

```{r}
#| label: product_checks

z |>
  mutate(era = cut(yr, c(1997, 2002, 2011, 2016, 2026), labels = c(
    "1998-2002 (SeaWiFS only)", "2003-2011 (+MODIS, MERIS)", "2012-2016 (+VIIRS)", "2017-2025 (+OLCI)"))) |>
  group_by(nms, era) |>
  summarize(CHL = round(mean(CHL), 2), .groups = "drop") |>
  pivot_wider(names_from = era, values_from = CHL) |>
  mutate(nms = factor(nms, NMS_ORD)) |>
  arrange(nms) |>
  kable(caption = "Mean standardized chlorophyll anomaly by ocean-color sensor era. A step shared by sanctuaries in different basins points to the merged product, not the ocean.")

z |>
  mutate(era = cut(yr, c(1997, 2002, 2006, 2013, 2016, 2021, 2026),
                   labels = c("1998-02", "2003-06", "2007-13", "2014-16", "2017-21", "2022-25"))) |>
  group_by(era) |>
  summarize(across(all_of(VARS), \(x) round(mean(x), 2)), .groups = "drop") |>
  kable(caption = "Mean standardized anomaly by period, averaged over all 13 sanctuaries")

z |>
  group_by(nms) |>
  summarize(
    `r(CHL, PP)`             = round(cor(CHL, PP), 2),
    `r(tempSurf, tempBott)`  = round(cor(tempSurf, tempBott), 2),
    .groups = "drop") |>
  mutate(nms = factor(nms, NMS_ORD)) |>
  arrange(nms) |>
  kable(caption = "Redundancy among response variables (monthly standardized anomalies)")

d_wide |>
  filter(date >= as_date("1998-01-01"), date <= as_date("2025-09-30")) |>
  mutate(mo = month(date)) |>
  group_by(nms, mo) |>
  summarize(across(all_of(VARS), \(x) sd(x, na.rm = T) < 1e-8), .groups = "drop") |>
  pivot_longer(all_of(VARS), names_to = "var") |>
  filter(value) |>
  select(nms, var, month = mo) |>
  kable(caption = "Sanctuary × calendar-month × variable cells with zero variance (e.g., mixed-layer depth pinned at the model's shallowest level)")
```

## Session

```{r}
#| label: session

sessionInfo()
```
