---
title: "Environmental trajectories of 13 National Marine Sanctuaries"
subtitle: "Departure from each sanctuary's 2003-2012 baseline, red-noise null, status and form, temporal eigenfunctions, common state space, events, climate modes and sensitivity"
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
    code-tools: true
execute:
  echo: true
  message: false
  warning: false
  error: false
editor_options:
  chunk_output_type: console
---

This notebook computes every number the trajectories manuscript reports and writes tidy CSVs to
`data/trajectories/`. Manuscript figures are drawn elsewhere from those CSVs; plots here are
diagnostics only (`figures/trajectories/diag_*.png`). Slow steps are cached to CSV and recomputed
whenever an input file is newer than its cache. Optional inputs (marine-heatwave days, climate
indices, OC-CCI chlorophyll) are used when present; their sections say so when they are skipped.

```{r}
#| label: setup

librarian::shelf(
  adespatial, dplyr, ecp, ggplot2, glue, here, knitr, lubridate, parallel,
  patchwork, purrr, readr, scales, stringr, tibble, tidyr, vegan, zoo,
  quiet = T)
options(readr.show_col_types = F, dplyr.summarise.inform = F)
set.seed(42)
t_start <- Sys.time()

# paths ----
state_csv   <- here("data/state/state_monthly.csv")
events_csv  <- here("data/events/events_catalog.csv")
mhw_csv     <- here("data/events/mhw_monthly.csv")
indices_csv <- here("data/context/indices_monthly.csv")
occci_csv   <- here("data/context/chl_occci_monthly.csv")
dir_out     <- here("data/trajectories")
dir_fig     <- here("figures/trajectories")
dir.create(dir_out, showWarnings = F, recursive = T)
dir.create(dir_fig, showWarnings = F, recursive = T)
out <- \(f) file.path(dir_out, f)

# constants ----
VARS      <- c("sst", "sbt", "sss", "mld", "chl", "precip")
LOG_VARS  <- c("chl", "mld", "pp", "chl_all", "chl_occci")
NMS_LUT   <- tribble(
  ~nms,      ~name,                             ~region,
  "OCNMS",   "Olympic Coast",                   "West Coast",
  "GFNMS",   "Greater Farallones",              "West Coast",
  "CBNMS",   "Cordell Bank",                    "West Coast",
  "MBNMS",   "Monterey Bay",                    "West Coast",
  "CPNMS",   "Chumash Heritage",                "West Coast",
  "CINMS",   "Channel Islands",                 "West Coast",
  "PNMS",    "Papahānaumokuākea",               "Pacific Islands",
  "HIHWNMS", "Hawaiian Islands Humpback Whale", "Pacific Islands",
  "NMSAS",   "American Samoa",                  "Pacific Islands",
  "FGBNMS",  "Flower Garden Banks",             "Southeast and Gulf",
  "FKNMS",   "Florida Keys",                    "Southeast and Gulf",
  "GRNMS",   "Gray's Reef",                     "Southeast and Gulf",
  "MNMS",    "Monitor",                         "Northeast",
  "SBNMS",   "Stellwagen Bank",                 "Northeast")
NMS_ALL   <- NMS_LUT$nms   # design order; NMS_ORD (set in the data chunk) = those present in the state file
BASIN     <- set_names(
  if_else(NMS_ALL %in% c("FGBNMS", "FKNMS", "GRNMS", "MNMS", "SBNMS"), "atlantic", "pacific"), NMS_ALL)
WIN_BEG   <- as_date("2003-01-01")
WIN_END   <- as_date("2025-09-01")
BASE_BEG  <- as_date("2003-01-01")
BASE_END  <- as_date("2012-12-01")
TEMP_BEG  <- as_date("1993-01-01")
WIN_DATES <- seq(WIN_BEG, WIN_END, by = "month")
N_WIN     <- length(WIN_DATES)
YR_WIN    <- year(WIN_DATES)
YEARS     <- sort(unique(YR_WIN))
IS_BASE   <- WIN_DATES >= BASE_BEG & WIN_DATES <= BASE_END
PERIODS   <- tribble(
  ~period,     ~y0,  ~y1,
  "base",      2003, 2012,
  "2013-2015", 2013, 2015,
  "2016-2020", 2016, 2020,
  "2021-2025", 2021, 2025,
  "post",      2013, 2025)
PERIODS_VAR <- bind_rows(PERIODS, tibble(period = "2016-2025", y0 = 2016, y1 = 2025))
in_period <- \(yr, p) { r <- PERIODS_VAR[PERIODS_VAR$period == p, ]; yr >= r$y0 & yr <= r$y1 }
PER_MASK  <- map(set_names(PERIODS_VAR$period), \(p) in_period(YR_WIN, p))
SD_FLOOR  <- 0.5
SIGMA_THR <- 2
N_SUR     <- 999          # red-noise null series per sanctuary
N_EV_SUR  <- 999          # circular shifts for the event coincidence test
N_MODE_SUR <- 499         # phase-randomized surrogates per climate index
N_AEM_NULL <- 39          # red-noise series per sanctuary for the eigenfunction reference (spec: 99; cut for run time)
NPERM_AEM  <- 999
NPERM_AEM_NULL <- 199
MAX_NA    <- 0.1          # a variable with more missing months than this (window or baseline) is dropped
N_CORE    <- max(1, detectCores() - 2)
BANDS     <- c("sub-annual", "interannual", "decadal", "trend-like")
INDICES   <- c("oni", "pdo", "npgo", "nao", "amo")   # natl_sst, nino34 etc. in the file are not used
LAGS      <- 0:6
# climate indices assigned a priori by basin
MODE_SETS <- list(pacific = c("oni", "pdo", "npgo"), atlantic = c("oni", "nao", "amo"))
SEASONS   <- c("DJF", "MAM", "JJA", "SON")
seed_of <- \(nms_i, offset = 0) 42 + 1000 * match(nms_i, NMS_ALL) + offset

# a cache is fresh when it exists and is newer than every input that is present ----
is_fresh <- function(f, inputs = state_csv) {
  inputs <- inputs[file.exists(inputs)]
  all(file.exists(f)) && all(outer(file.mtime(f), file.mtime(inputs), ">"))
}
note <- \(...) cat(glue("NOTE: ", ..., .envir = parent.frame()), "\n")

# plot theme (from trajectory_checks.qmd) ----
PAL <- list(
  surface = "#fcfcfb", ink = "#0b0b0b", ink2 = "#52514e", muted = "#898781",
  grid    = "#e1e0d9", axis = "#c3c2b7", blue = "#2a78d6", orange = "#eb6834")
PAL_DIV  <- c("#104281", "#2a78d6", "#9ec5f4", "#f0efec", "#f3b3ad", "#e34948", "#8f1d1c")
PAL_BAND <- c("#86b6ef", "#3987e5", "#1c5cab", "#0d366b")   # ordinal, short -> long time scale

theme_chk <- function(base_size = 10) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.background    = element_rect(fill = PAL$surface, color = NA),
      panel.background   = element_rect(fill = PAL$surface, color = NA),
      panel.grid.major   = element_line(color = PAL$grid, linewidth = 0.25),
      panel.grid.minor   = element_blank(),
      axis.text          = element_text(color = PAL$muted),
      axis.title         = element_text(color = PAL$ink2),
      plot.title         = element_text(color = PAL$ink, face = "bold", size = rel(1.1)),
      plot.subtitle      = element_text(color = PAL$ink2),
      plot.caption       = element_text(color = PAL$muted, hjust = 0),
      plot.title.position = "plot",
      strip.text         = element_text(color = PAL$ink, face = "bold", hjust = 0),
      legend.position    = "top",
      legend.justification = "left",
      legend.title       = element_text(color = PAL$ink2),
      legend.text        = element_text(color = PAL$ink2))
}
save_fig <- function(p, name, width = 9, height = 5) {
  ggsave(file.path(dir_fig, name), p, width = width, height = height, dpi = 200, bg = PAL$surface)
  p
}
```

## Data

```{r}
#| label: data

d_state <- read_csv(state_csv) |>
  mutate(date = as_date(date)) |>
  filter(nms %in% NMS_ALL)
NMS_ORD <- intersect(NMS_ALL, unique(d_state$nms))
NMS_LUT <- NMS_LUT |> filter(nms %in% NMS_ORD)
if (length(setdiff(NMS_ALL, NMS_ORD)) > 0)
  note("sanctuaries absent from the state file, not analyzed: {paste(setdiff(NMS_ALL, NMS_ORD), collapse = ', ')}")
stopifnot(
  all(c("nms", "date", "var", "value") %in% names(d_state)),
  all(VARS %in% d_state$var),
  !anyDuplicated(d_state[, c("nms", "date", "var")]))
is_provisional <- "rule" %in% names(d_state) && any(str_detect(coalesce(d_state$rule, ""), "provisional"))

inputs <- tibble(
  input = c("state_monthly", "events_catalog", "mhw_monthly", "indices_monthly", "chl_occci_monthly"),
  path  = c(state_csv, events_csv, mhw_csv, indices_csv, occci_csv)) |>
  mutate(
    present  = file.exists(path),
    modified = if_else(present, format(file.mtime(path), "%Y-%m-%d %H:%M"), NA_character_),
    note     = case_when(
      input == "state_monthly" & is_provisional ~ "provisional (old unmasked extracts)",
      input == "state_monthly"                  ~ "final",
      !present                                  ~ "absent: section skipped",
      TRUE                                      ~ "present"),
    path = str_remove(path, paste0(here(), "/")))
kable(inputs, caption = "Inputs at this render")

d_state |>
  filter(var %in% c(VARS, "pp", "sbt_all", "chl_all"), date <= WIN_END) |>
  group_by(var) |>
  summarize(
    first = min(date[!is.na(value)]), last = max(date[!is.na(value)]),
    sanctuaries = n_distinct(nms[!is.na(value)]), missing = sum(is.na(value))) |>
  kable(caption = "Coverage of the state variables (up to the end of the window)")
```

## Helper functions

```{r}
#| label: functions

# transforms ----
transform_x <- function(var, value) {
  out <- value
  i   <- var %in% LOG_VARS
  out[i] <- ifelse(value[i] > 0, log10(pmax(value[i], 1e-30)), NA_real_)
  j   <- var == "precip"
  out[j] <- sqrt(pmax(value[j], 0))
  out
}

# calendar-month standardization against a baseline, with an sd floor ----
standardize <- function(d, base_beg = BASE_BEG, base_end = BASE_END) {
  d |>
    mutate(
      x  = transform_x(var, value),
      mo = month(date),
      b  = date >= base_beg & date <= base_end & is.finite(x)) |>
    group_by(nms, var, mo) |>
    mutate(
      mu_m = if (any(b)) mean(x[b]) else NA_real_,
      sd_m = if (sum(b) > 1) sd(x[b]) else NA_real_) |>
    group_by(nms, var) |>
    mutate(
      sd_pooled = if (sum(b) > 2) sd((x - mu_m)[b]) else NA_real_,
      sd_use    = pmax(sd_m, SD_FLOOR * sd_pooled),
      floor_hit = sd_m < SD_FLOOR * sd_pooled,
      z         = ifelse(is.finite(x) & is.finite(sd_use) & sd_use > 0, (x - mu_m) / sd_use, NA_real_)) |>
    ungroup()
}

# per-sanctuary matrix of z over the window; drop variables with too many gaps ----
build_Z <- function(dz, nms_i, vars = VARS, is_base = IS_BASE, dates = WIN_DATES) {
  w <- dz |>
    filter(nms == nms_i, var %in% vars) |>
    select(date, var, z) |>
    pivot_wider(names_from = var, values_from = z)
  w <- tibble(date = dates) |> left_join(w, by = "date")
  v <- intersect(vars, names(w))
  Z <- as.matrix(w[, v, drop = FALSE])
  Z[!is.finite(Z)] <- NA
  ok <- colMeans(!is.na(Z)) >= 1 - MAX_NA & colMeans(!is.na(Z[is_base, , drop = FALSE])) >= 1 - MAX_NA
  list(
    nms = nms_i, Z = Z[, ok, drop = FALSE], vars = v[ok],
    dropped = setdiff(vars, v[ok]), n_na = sum(is.na(Z[, ok])))
}
na0 <- \(Z) { Z[is.na(Z)] <- 0; Z }

# sigma dissimilarity (mahony et al. 2017) ----
sigma_from_d2 <- function(d2, df) qnorm(pmin(pchisq(d2, df = df), 1 - 1e-12) / 2 + 0.5)

# mahalanobis distance; rows with gaps use the available variables (df = their number)
mahal_na <- function(Z, center, S) {
  d2 <- rep(NA_real_, nrow(Z))
  df <- rep(ncol(Z), nrow(Z))
  ok <- complete.cases(Z)
  if (any(ok)) d2[ok] <- mahalanobis(Z[ok, , drop = FALSE], center, S)
  for (i in which(!ok)) {
    j <- !is.na(Z[i, ])
    if (any(j)) {
      d2[i] <- mahalanobis(Z[i, j, drop = FALSE], center[j], S[j, j, drop = FALSE])
      df[i] <- sum(j)
    }
  }
  list(d2 = d2, df = df)
}

# departure from the baseline envelope; baseline months scored leave-one-year-out ----
score_departure <- function(Z, is_base = IS_BASE, yr = YR_WIN) {
  B     <- Z[is_base, , drop = FALSE]
  m     <- mahal_na(Z, colMeans(B, na.rm = TRUE), cov(B, use = "pairwise.complete.obs"))
  sigma <- sigma_from_d2(m$d2, m$df)
  sigma_cv <- sigma
  for (y in unique(yr[is_base])) {
    tr <- is_base & yr != y
    te <- is_base & yr == y
    Bt <- Z[tr, , drop = FALSE]
    mm <- mahal_na(Z[te, , drop = FALSE], colMeans(Bt, na.rm = TRUE), cov(Bt, use = "pairwise.complete.obs"))
    sigma_cv[te] <- sigma_from_d2(mm$d2, mm$df)
  }
  list(sigma = sigma, sigma_cv = sigma_cv)
}

# every statistic used for status / trend / form, computed identically for data and null ----
dep_stats <- function(Z, thr = SIGMA_THR, is_base = IS_BASE) {
  sc    <- score_departure(Z, is_base, YR_WIN)
  ex    <- sc$sigma_cv > thr
  e_per <- vapply(PERIODS$period, \(p) 100 * mean(ex[PER_MASK[[p]]], na.rm = TRUE), numeric(1))
  e_ann <- vapply(YEARS, \(y) 100 * mean(ex[YR_WIN == y], na.rm = TRUE), numeric(1))
  run3  <- as.numeric(stats::filter(e_ann, rep(1 / 3, 3), sides = 1))
  v_sh  <- vapply(
    PERIODS_VAR$period,
    \(p) colMeans(abs(Z[PER_MASK[[p]], , drop = FALSE]) > 2, na.rm = TRUE), numeric(ncol(Z)))
  v_sh  <- matrix(v_sh, nrow = ncol(Z), dimnames = list(colnames(Z), PERIODS_VAR$period))
  v_mn  <- vapply(
    PERIODS_VAR$period,
    \(p) colMeans(Z[PER_MASK[[p]], , drop = FALSE], na.rm = TRUE), numeric(ncol(Z)))
  v_mn  <- matrix(v_mn, nrow = ncol(Z), dimnames = list(colnames(Z), PERIODS_VAR$period))
  list(
    sigma = sc$sigma, sigma_cv = sc$sigma_cv, exceed = ex,
    e_per = e_per, e_ann = setNames(e_ann, YEARS), run3 = setNames(run3, YEARS), v_sh = v_sh, v_mn = v_mn)
}

# var(1) red noise ----
fit_var1 <- function(Zb) {
  X   <- Zb[-nrow(Zb), , drop = FALSE]
  Y   <- Zb[-1, , drop = FALSE]
  Bm  <- solve(crossprod(X), crossprod(X, Y))      # Y = X Bm, so A = t(Bm)
  A   <- t(Bm)
  E   <- Y - X %*% Bm
  Sig <- crossprod(E) / (nrow(E) - ncol(E))
  eig <- max(Mod(eigen(A, only.values = TRUE)$values))
  shrunk <- eig >= 1
  if (shrunk) A <- A * 0.98 / eig
  list(A = A, Sig = Sig, eig_max = eig, shrunk = shrunk, vars = colnames(Zb))
}
sim_var1 <- function(fit, n = N_WIN, burn = 120) {
  p <- ncol(fit$A)
  L <- tryCatch(chol(fit$Sig), error = \(e) chol(fit$Sig + diag(1e-8, p)))
  e <- matrix(rnorm((n + burn) * p), ncol = p) %*% L
  S <- matrix(0, n + burn, p)
  for (i in 2:(n + burn)) S[i, ] <- fit$A %*% S[i - 1, ] + e[i, ]
  S <- S[-seq_len(burn), , drop = FALSE]
  colnames(S) <- fit$vars
  S
}
# re-standardize each variable by the mean and sd of the first n_b months
restd <- function(S, n_b = sum(IS_BASE)) {
  mu <- colMeans(S[seq_len(n_b), , drop = FALSE])
  s  <- apply(S[seq_len(n_b), , drop = FALSE], 2, sd)
  sweep(sweep(S, 2, mu), 2, s, "/")
}
p_emp <- \(null, obs) (1 + sum(null >= obs - 1e-9)) / (1 + length(null))

# temporal eigenfunctions (aem) with positive moran's i (helper from trajectory_checks.qmd) ----
aem_cache <- new.env()
get_aem <- function(n) {
  key <- as.character(n)
  if (is.null(aem_cache[[key]])) {
    set.seed(1)
    a  <- aem.time(n, moran = TRUE)
    mi <- as.numeric(a$Moran$obs)
    E  <- a$aem
    colnames(E) <- paste0("AEM", seq_len(ncol(E)))
    aem_cache[[key]] <- E[, mi > 0, drop = FALSE]
  }
  aem_cache[[key]]
}
period_of_k <- \(k, n = N_WIN) 2 * (n / 12) / k
band_of     <- \(period_yr) as.character(cut(period_yr, c(0, 1.5, 8, 15, Inf), labels = BANDS))

# forward selection with the double stopping rule, as in the pilot's full_series chunk
fit_aem_sel <- function(Y, E, nperm = NPERM_AEM, alpha = 0.05) {
  none <- tibble(k = integer(), r2adj_increment = numeric(), p_value = numeric())
  r2g  <- RsquareAdj(rda(Y, E))$adj.r.squared
  if (!is.finite(r2g) || r2g <= 0) return(none)
  fs <- tryCatch(
    forward.sel(Y, E, alpha = alpha, adjR2thresh = r2g, nperm = nperm, verbose = FALSE),
    error = \(e) NULL)
  if (is.null(fs) || nrow(fs) == 0) return(none)
  tibble(
    k               = as.integer(str_remove(colnames(E)[fs$order], "AEM")),
    r2adj_increment = c(fs$AdjR2Cum[1], diff(fs$AdjR2Cum)),
    p_value         = fs$pvalue)
}
# rda on selected eigenfunctions (get_rda in the pilot); fitted states by least squares
get_rda <- \(Y, E, ks) rda(Y, E[, paste0("AEM", ks), drop = FALSE])
fit_states <- function(Y, E, ks) {
  X <- cbind(1, E[, paste0("AEM", ks), drop = FALSE])
  X %*% qr.coef(qr(X), Y)
}

# adjusted r2 of a multivariate linear model (= RsquareAdj(rda(Y, X))) ----
r2adj_fast <- function(Y, X) {
  X  <- as.matrix(X)
  ok <- complete.cases(X)
  Y  <- Y[ok, , drop = FALSE]; X <- X[ok, , drop = FALSE]
  n  <- nrow(Y); m <- ncol(X)
  Yc <- sweep(Y, 2, colMeans(Y))
  Xc <- sweep(X, 2, colMeans(X))
  if (m == 1) {
    r2 <- sum(crossprod(Xc, Yc)^2) / (sum(Xc^2) * sum(Yc^2))
  } else {
    r2 <- sum(qr.fitted(qr(Xc), Yc)^2) / sum(Yc^2)
  }
  1 - (1 - r2) * (n - 1) / (n - m - 1)
}

# phase randomization of one series (keeps its power spectrum) ----
phase_rand <- function(v) {
  ok  <- which(!is.na(v)); seg <- min(ok):max(ok)
  x   <- v[seg]
  if (anyNA(x)) x <- zoo::na.approx(x)
  n   <- length(x); m <- mean(x)
  k   <- floor((n - 1) / 2)
  ph  <- runif(k, 0, 2 * pi)
  rot <- rep(1 + 0i, n)
  rot[2:(k + 1)]     <- exp(1i * ph)
  rot[n:(n - k + 1)] <- exp(-1i * ph)
  out <- v
  out[seg] <- Re(fft(fft(x - m) * rot, inverse = TRUE)) / n + m
  out
}
# circular shift of a series by o positions
circ_shift <- \(x, o) x[((seq_along(x) - 1 + o) %% length(x)) + 1]
```

### Checks of the helpers on synthetic data

```{r}
#| label: helper_checks

set.seed(7)
chk <- list()

# sigma of a point 2 sd out on one of p independent variables (identity covariance)
chk$sigma <- tibble(p = c(1, 2, 6)) |>
  mutate(
    `2 SD on one variable`   = sigma_from_d2(4, p),
    `2 SD on every variable` = sigma_from_d2(4 * p, p))
stopifnot(
  abs(chk$sigma$`2 SD on one variable`[1] - 2) < 1e-9,
  abs(sigma_from_d2(4, 6) - qnorm(pchisq(4, 6) / 2 + 0.5)) < 1e-12,
  all(diff(chk$sigma$`2 SD on one variable`) < 0))

# white noise: exceedance near 5%; a 1.5 sd step in one of six variables is detected
wn <- map_dfr(1:200, \(r) {
  S  <- restd(matrix(rnorm(N_WIN * 6), ncol = 6, dimnames = list(NULL, VARS)))
  st <- dep_stats(S)
  S2 <- S; S2[!IS_BASE, 1] <- S2[!IS_BASE, 1] + 1.5
  st2 <- dep_stats(S2)
  tibble(base_cv = st$e_per["base"], post = st$e_per["post"], post_step = st2$e_per["post"],
         base_in = 100 * mean(st$sigma[IS_BASE] > 2))
})
chk$wn <- wn |> summarize(across(everything(), mean))
stopifnot(
  between(chk$wn$base_in, 3, 7),
  between(chk$wn$base_cv, 3, 12),
  between(chk$wn$post, 3, 10),
  chk$wn$post_step > 15)

# var(1) recovery on a long series
A_true <- matrix(c(0.7, 0.1, 0, 0.5), 2)
f_true <- list(A = A_true, Sig = diag(2), vars = c("a", "b"))
f_hat  <- fit_var1(sim_var1(f_true, n = 20000))
stopifnot(max(abs(f_hat$A - A_true)) < 0.03, max(abs(f_hat$Sig - diag(2))) < 0.05)

# sd floor: a month with constant values gets finite z through the floor
d_syn <- tibble(date = seq(as_date("2003-01-01"), as_date("2012-12-01"), by = "month")) |>
  mutate(nms = "X", var = "sst", value = rnorm(n()) + if_else(month(date) == 7, 0, 0))
d_syn$value[month(d_syn$date) == 7] <- 10
z_syn <- standardize(d_syn)
stopifnot(
  all(z_syn$floor_hit[month(z_syn$date) == 7]),
  all(is.finite(z_syn$z)),
  !any(z_syn$floor_hit[month(z_syn$date) != 7]))

# r2adj_fast equals vegan's adjusted r2 of the rda
Yt <- matrix(rnorm(300), 50); Xt <- matrix(rnorm(100), 50)
stopifnot(
  abs(r2adj_fast(Yt, Xt) - RsquareAdj(rda(Yt, Xt))$adj.r.squared) < 1e-10,
  abs(r2adj_fast(Yt, Xt[, 1, drop = FALSE]) - RsquareAdj(rda(Yt, Xt[, 1, drop = FALSE]))$adj.r.squared) < 1e-10)

# phase randomization keeps the amplitude spectrum; circular shift keeps the count
xs <- cumsum(rnorm(285)); xr <- phase_rand(xs)
stopifnot(
  max(abs(Mod(fft(xs - mean(xs))) - Mod(fft(xr - mean(xr))))) < 1e-6,
  sum(circ_shift(1:10 > 7, 4)) == 3)

# time-scale bands
stopifnot(
  band_of(period_of_k(3)) == "trend-like", band_of(period_of_k(4)) == "decadal",
  band_of(period_of_k(30)) == "interannual", band_of(period_of_k(31)) == "sub-annual")

chk$sigma |> mutate(across(-p, \(x) round(x, 2))) |>
  kable(caption = "Sigma dissimilarity of a point 2 SD from the centre (identity covariance): one variable vs all p")
chk$wn |> mutate(across(everything(), \(x) round(x, 1))) |>
  rename(`baseline, in sample (%)` = base_in, `baseline, leave-one-year-out (%)` = base_cv,
         `2013-2025 (%)` = post, `2013-2025 with a 1.5 SD step in one variable (%)` = post_step) |>
  kable(caption = "Share of months beyond 2 sigma for white noise (p = 6, 200 series)")
```

## 1. Standardized anomalies

Each variable is transformed (log10 for chlorophyll and mixed-layer depth, square root for
precipitation), then standardized by calendar month against 2003-2012. The SD of a calendar month
is floored at half the pooled baseline SD, so near-constant months do not inflate anomalies.

```{r}
#| label: z

dz_all <- standardize(d_state |> filter(var %in% VARS))
dz     <- dz_all |> filter(date >= WIN_BEG, date <= WIN_END)

Z_list <- map(set_names(NMS_ORD), \(n) build_Z(dz, n))
stopifnot(all(map_int(Z_list, \(zi) nrow(zi$Z)) == N_WIN), all(map_int(Z_list, \(zi) ncol(zi$Z)) >= 2))
P_OF <- map_int(Z_list, \(zi) ncol(zi$Z))

dz |>
  select(nms, date, var, value, x, z) |>
  arrange(factor(nms, NMS_ORD), var, date) |>
  write_csv(out("z_monthly.csv"))
dz_all |>
  filter(var %in% c("sst", "sbt"), date >= TEMP_BEG, date <= WIN_END) |>
  select(nms, date, var, value, x, z) |>
  arrange(factor(nms, NMS_ORD), var, date) |>
  write_csv(out("z_temp_1993.csv"))

floor_tbl <- dz |>
  distinct(nms, var, mo, floor_hit) |>
  group_by(nms, var) |>
  summarize(n = sum(floor_hit, na.rm = TRUE)) |>
  pivot_wider(names_from = var, values_from = n) |>
  mutate(nms = factor(nms, NMS_ORD)) |>
  arrange(nms)
n_floor_cells  <- dz |> distinct(nms, var, mo, floor_hit) |> summarize(n = sum(floor_hit, na.rm = TRUE)) |> pull(n)
n_floor_months <- dz |> summarize(n = sum(floor_hit & !is.na(z), na.rm = TRUE)) |> pull(n)
floor_tbl |>
  kable(caption = glue(
    "Calendar months (of 12) whose SD was raised to the floor, by sanctuary and variable. ",
    "In total {n_floor_cells} sanctuary-variable-calendar-month cells, i.e. {n_floor_months} ",
    "sanctuary-variable-months in the window."))

tibble(
  nms     = NMS_ORD,
  p       = P_OF,
  used    = map_chr(Z_list, \(zi) paste(zi$vars, collapse = ", ")),
  dropped = map_chr(Z_list, \(zi) paste(zi$dropped, collapse = ", ")),
  `missing z filled for models` = map_int(Z_list, \(zi) zi$n_na)) |>
  kable(caption = "Variables used per sanctuary")
```

## 2. Multivariate departure from the baseline envelope

For each month, the Mahalanobis distance of the anomaly vector from the baseline centre (baseline
covariance) is expressed as sigma dissimilarity, the equivalent number of SDs of a univariate
normal. Baseline months are scored leave-one-year-out (`sigma_cv`).

```{r}
#| label: departure

obs_stats <- map(Z_list, \(zi) dep_stats(zi$Z))
d_dep <- imap_dfr(obs_stats, \(st, n) tibble(
  nms      = n,
  date     = WIN_DATES,
  sigma    = st$sigma,
  sigma_cv = st$sigma_cv,
  exceed   = st$exceed,
  d_rms    = sqrt(rowMeans(Z_list[[n]]$Z^2, na.rm = TRUE)))) |>
  group_by(nms) |>
  mutate(
    sigma_run12  = rollmean(sigma_cv, 12, fill = NA, align = "center"),
    exceed_run12 = rollmean(as.numeric(exceed), 12, fill = NA, align = "center")) |>
  ungroup()
stopifnot(nrow(d_dep) == N_WIN * length(NMS_ORD), !anyNA(d_dep$sigma_cv))

e_base <- map_dbl(obs_stats, \(st) st$e_per["base"])
e_base_in <- map_dbl(obs_stats, \(st) 100 * mean(st$sigma[IS_BASE] > SIGMA_THR))
stopifnot(all(between(e_base, 1, 20)))
if (any(!between(e_base, 4, 13)))
  note("baseline out-of-sample exceedance outside 4-13% for {paste(NMS_ORD[!between(e_base, 4, 13)], collapse = ', ')}")

d_dep |>
  mutate(nms = factor(nms, NMS_ORD), yr = year(date)) |>
  group_by(nms) |>
  summarize(
    `baseline in sample` = 100 * mean(sigma[yr <= 2012] > SIGMA_THR),
    `baseline out of sample` = 100 * mean(exceed[yr <= 2012]),
    `2013-2015` = 100 * mean(exceed[between(yr, 2013, 2015)]),
    `2016-2020` = 100 * mean(exceed[between(yr, 2016, 2020)]),
    `2021-2025` = 100 * mean(exceed[yr >= 2021])) |>
  mutate(across(-nms, \(x) round(x, 1))) |>
  kable(caption = glue("Share of months (%) beyond {SIGMA_THR} sigma"))
```

```{r}
#| label: fig_sigma
#| fig-width: 10
#| fig-height: 8

p <- d_dep |>
  mutate(nms = factor(nms, NMS_ORD)) |>
  ggplot(aes(date, sigma_cv)) +
  annotate("rect", xmin = BASE_BEG, xmax = BASE_END, ymin = -Inf, ymax = Inf, fill = PAL$grid, alpha = 0.5) +
  geom_hline(yintercept = SIGMA_THR, color = PAL$orange, linewidth = 0.3) +
  geom_line(color = PAL$axis, linewidth = 0.25) +
  geom_line(aes(y = sigma_run12), color = PAL$blue, linewidth = 0.6) +
  facet_wrap(~nms, ncol = 3) +
  labs(
    title = "Departure from the 2003-2012 envelope",
    subtitle = "Monthly sigma dissimilarity (grey; baseline scored leave-one-year-out), 12-month running mean (blue), 2 sigma (orange)",
    x = NULL, y = "sigma") +
  theme_chk()
save_fig(p, "diag_sigma.png", 10, 8)
```

## 3. Null model: stationary red noise with the baseline's own persistence

A VAR(1) fitted to each sanctuary's baseline anomalies generates `r N_SUR` stationary series of
`r N_WIN` months. Each is re-standardized on its first 120 months and scored exactly as the data,
which gives the exceedance expected from persistence alone, without any change.

```{r}
#| label: null_model

var1_fits <- map(Z_list, \(zi) fit_var1(na0(zi$Z)[IS_BASE, , drop = FALSE]))
null_files <- out(c("null_draws.csv", "null_exceedance.csv", "null_var.csv"))
# recompute when an input is newer, or when the cache predates the period-mean quantiles
null_ok <- is_fresh(null_files) &&
  all(c("q025_mean", "q975_mean") %in% names(read_csv(null_files[3], n_max = 1))) &&
  setequal(unique(read_csv(null_files[2])$nms), NMS_ORD)

if (!null_ok) {
  run_null <- function(nms_i) {
    set.seed(seed_of(nms_i))
    fit  <- var1_fits[[nms_i]]
    p    <- ncol(fit$A)
    e    <- matrix(NA_real_, N_SUR, nrow(PERIODS), dimnames = list(NULL, PERIODS$period))
    vs   <- array(NA_real_, c(N_SUR, p, nrow(PERIODS_VAR)), dimnames = list(NULL, fit$vars, PERIODS_VAR$period))
    vm   <- vs
    for (r in seq_len(N_SUR)) {
      st <- dep_stats(restd(sim_var1(fit)))
      e[r, ]    <- st$e_per
      vs[r, , ] <- st$v_sh
      vm[r, , ] <- st$v_mn
    }
    list(
      draws  = as_tibble(e) |> mutate(nms = nms_i, rep = row_number()) |>
        pivot_longer(-c(nms, rep), names_to = "period", values_to = "e_pct"),
      var    = expand_grid(var = fit$vars, period = PERIODS_VAR$period) |>
        mutate(
          nms       = nms_i,
          q95_share = map2_dbl(var, period, \(v, pp) quantile(vs[, v, pp], 0.95)),
          q025_mean = map2_dbl(var, period, \(v, pp) quantile(vm[, v, pp], 0.025)),
          q975_mean = map2_dbl(var, period, \(v, pp) quantile(vm[, v, pp], 0.975))) |>
        select(nms, var, period, q95_share, q025_mean, q975_mean))
  }
  res <- mclapply(NMS_ORD, run_null, mc.cores = N_CORE)
  stopifnot(!any(map_lgl(res, \(r) inherits(r, "try-error"))))
  d_null_draws <- map_dfr(res, "draws")
  write_csv(d_null_draws, null_files[1])
  d_null_draws |>
    group_by(nms, period) |>
    summarize(q50 = quantile(e_pct, 0.5), q95 = quantile(e_pct, 0.95), q99 = quantile(e_pct, 0.99)) |>
    ungroup() |>
    write_csv(null_files[2])
  write_csv(map_dfr(res, "var"), null_files[3])
}
d_null_draws <- read_csv(null_files[1])
d_null_e     <- read_csv(null_files[2])
d_null_var   <- read_csv(null_files[3])
stopifnot(
  nrow(d_null_draws) == N_SUR * nrow(PERIODS) * length(NMS_ORD),
  all(between(d_null_e$q50[d_null_e$period == "base"], 2, 12)),
  all(between(d_null_e$q50[d_null_e$period == "post"], 1, 15)))

tibble(
  nms       = NMS_ORD,
  p         = P_OF,
  `max |eigenvalue| of A` = round(map_dbl(var1_fits, "eig_max"), 3),
  shrunk    = map_lgl(var1_fits, "shrunk")) |>
  left_join(
    d_null_e |>
      mutate(q = glue("{round(q50, 1)} / {round(q95, 1)}")) |>
      select(nms, period, q) |>
      pivot_wider(names_from = period, values_from = q),
    by = "nms") |>
  kable(caption = "VAR(1) persistence and null exceedance (%), median / 95th percentile, by period")
```

## 4. Status, trend and form

Each period's exceedance is compared with its own null 95th percentile. Status: outside when
2021-2025 exceeds its null; trend: change since 2016-2020 (10 points); form: the shape of the
three post-baseline periods; first period outside: the first of the three whose exceedance is above its null 95th percentile.

```{r}
#| label: status

d_eper <- imap_dfr(obs_stats, \(st, n) tibble(
  nms          = n,
  period       = PERIODS$period,
  e_pct        = unname(st$e_per),
  sigma_median = map_dbl(PERIODS$period, \(p) median(st$sigma_cv[PER_MASK[[p]]], na.rm = TRUE)))) |>
  left_join(d_null_e |> select(nms, period, null_q50 = q50, null_q95 = q95), by = c("nms", "period")) |>
  rowwise() |>
  mutate(p_value = p_emp(d_null_draws$e_pct[d_null_draws$nms == nms & d_null_draws$period == period], e_pct)) |>
  ungroup()

d_eann <- imap_dfr(obs_stats, \(st, n) tibble(
  nms        = n,
  year       = YEARS,
  e_pct      = unname(st$e_ann),
  sigma_mean = map_dbl(YEARS, \(y) mean(st$sigma_cv[YR_WIN == y], na.rm = TRUE)),
  n_months   = map_int(YEARS, \(y) sum(!is.na(st$sigma_cv[YR_WIN == y])))))
d_eann |> select(nms, year, e_pct, sigma_mean, n_months) |> write_csv(out("exceedance_annual.csv"))

d_shift <- imap_dfr(Z_list, \(zi, n) expand_grid(var = zi$vars, period = c("2016-2025", "2021-2025")) |>
  mutate(
    nms      = n,
    mean_z   = map2_dbl(var, period, \(v, p) mean(zi$Z[PER_MASK[[p]], v], na.rm = TRUE)),
    share_hi = map2_dbl(var, period, \(v, p) mean(zi$Z[PER_MASK[[p]], v] > 2, na.rm = TRUE)),
    share_lo = map2_dbl(var, period, \(v, p) mean(zi$Z[PER_MASK[[p]], v] < -2, na.rm = TRUE)))) |>
  left_join(
    d_null_var |> rename(null_q95_share = q95_share, null_q025_mean = q025_mean, null_q975_mean = q975_mean),
    by = c("nms", "var", "period")) |>
  mutate(
    sig      = share_hi + share_lo > null_q95_share,
    sig_mean = mean_z < null_q025_mean | mean_z > null_q975_mean) |>
  select(nms, var, period, mean_z, share_hi, share_lo, null_q95_share, sig, null_q025_mean, null_q975_mean, sig_mean)
stopifnot(!anyNA(d_shift$null_q95_share), !anyNA(d_shift$sig_mean))
write_csv(d_shift, out("shift_by_var.csv"))

# status, trend, form, first period outside ----
classify_form <- function(E, Q) {
  x <- E > Q
  m <- max(E[1:2])
  case_when(
    !x[3] & !any(x[1:2]) ~ "within baseline range",
    !x[3]                ~ "excursion and return",
    E[3] >= m + 10       ~ "growing departure",
    E[3] <= m - 10       ~ "partial return",
    TRUE                 ~ "sustained departure")
}
d_status <- d_eper |>
  filter(period %in% c("2013-2015", "2016-2020", "2021-2025")) |>
  group_by(nms) |>
  arrange(period, .by_group = TRUE) |>
  summarize(
    status = if_else(e_pct[3] > null_q95[3], "outside", "within"),
    delta  = e_pct[3] - e_pct[2],
    trend  = case_when(delta >= 10 ~ "increasing", delta <= -10 ~ "decreasing", TRUE ~ "steady"),
    form   = classify_form(e_pct, null_q95),
    first_period_outside = if (any(e_pct > null_q95)) period[which(e_pct > null_q95)[1]] else NA_character_)
stopifnot(nrow(d_status) == length(NMS_ORD))

d_eper |>
  filter(period != "post") |>
  mutate(cell = glue("{round(e_pct)} ({round(null_q95)})")) |>
  select(nms, period, cell) |>
  pivot_wider(names_from = period, values_from = cell) |>
  left_join(d_status |> mutate(delta = round(delta)), by = "nms") |>
  mutate(nms = factor(nms, NMS_ORD)) |>
  arrange(nms) |>
  kable(caption = "Exceedance (%) by period with the null 95th percentile in brackets; status, trend and form")

d_shift |>
  filter(period == "2021-2025") |>
  mutate(cell = paste0(sprintf("%+.1f", mean_z), if_else(sig_mean, "†", ""), if_else(sig, "*", ""))) |>
  select(nms, var, cell) |>
  pivot_wider(names_from = var, values_from = cell) |>
  mutate(nms = factor(nms, NMS_ORD)) |>
  arrange(nms) |>
  select(nms, any_of(VARS)) |>
  kable(caption = "Mean z in 2021-2025 by variable († = mean outside the null 2.5-97.5 percentile range; * = share of months with |z| > 2 above the null 95th percentile)")
```

```{r}
#| label: fig_null
#| fig-width: 9
#| fig-height: 5

p <- d_eper |>
  filter(period %in% c("base", "2013-2015", "2016-2020", "2021-2025")) |>
  mutate(nms = factor(nms, rev(NMS_ORD)), period = factor(period, c("base", "2013-2015", "2016-2020", "2021-2025"))) |>
  ggplot(aes(y = nms)) +
  geom_linerange(aes(xmin = null_q50, xmax = null_q95), color = PAL$axis, linewidth = 2) +
  geom_point(aes(x = e_pct), color = PAL$orange, size = 1.8) +
  facet_wrap(~period, nrow = 1) +
  labs(
    title = "Observed exceedance vs the red-noise null",
    subtitle = "Share of months beyond 2 sigma (orange); grey bar = null median to 95th percentile",
    x = "% of months", y = NULL) +
  theme_chk()
save_fig(p, "diag_null.png", 9, 5)
```

## 5. Temporal eigenfunction model (AEM-RDA)

One model per sanctuary on all `r N_WIN` months: eigenfunctions with positive Moran's I, forward
selection with the double stopping rule, then RDA. Selected eigenfunctions are grouped by full-cycle
period into four time-scale bands.

```{r}
#| label: aem

E_aem <- get_aem(N_WIN)
stopifnot(nrow(E_aem) == N_WIN, ncol(E_aem) > 50)
aem_files <- out(c("aem_selected.csv"))

if (!is_fresh(aem_files)) {
  res <- mclapply(NMS_ORD, \(n) {
    set.seed(seed_of(n, 1))
    fit_aem_sel(na0(Z_list[[n]]$Z), E_aem, nperm = NPERM_AEM) |> mutate(nms = n, .before = 1)
  }, mc.cores = N_CORE)
  stopifnot(!any(map_lgl(res, \(r) inherits(r, "try-error"))))
  bind_rows(res) |>
    mutate(period_yr = period_of_k(k), band = band_of(period_yr)) |>
    select(nms, k, period_yr, band, r2adj_increment, p_value) |>
    write_csv(aem_files[1])
}
d_aem_sel <- read_csv(aem_files[1])

d_aem_bands <- expand_grid(nms = NMS_ORD, band = BANDS) |>
  left_join(d_aem_sel |> group_by(nms, band) |> summarize(r2adj = sum(r2adj_increment)), by = c("nms", "band")) |>
  mutate(r2adj = replace_na(r2adj, 0))
d_aem_tot <- tibble(nms = NMS_ORD, p = P_OF) |>
  left_join(d_aem_sel |> group_by(nms) |> summarize(r2adj = sum(r2adj_increment), n_selected = n()), by = "nms") |>
  mutate(r2adj = replace_na(r2adj, 0), n_selected = replace_na(n_selected, 0L)) |>
  select(nms, r2adj, n_selected, p)
write_csv(d_aem_bands, out("aem_bands.csv"))
write_csv(d_aem_tot, out("aem_total.csv"))

# fitted states; the rda adjusted r2 must equal the summed increments
d_fit <- map_dfr(NMS_ORD, \(n) {
  Y  <- na0(Z_list[[n]]$Z)
  ks <- d_aem_sel$k[d_aem_sel$nms == n]
  if (length(ks) > 0) {
    r2 <- RsquareAdj(get_rda(Y, E_aem, ks))$adj.r.squared
    stopifnot(abs(r2 - sum(d_aem_sel$r2adj_increment[d_aem_sel$nms == n])) < 1e-6)
    Fz <- fit_states(Y, E_aem, ks)
  } else {
    Fz <- matrix(colMeans(Y), N_WIN, ncol(Y), byrow = TRUE, dimnames = list(NULL, colnames(Y)))
  }
  as_tibble(Fz) |> mutate(nms = n, date = WIN_DATES) |>
    pivot_longer(-c(nms, date), names_to = "var", values_to = "z_fit")
}) |> select(nms, date, var, z_fit)
write_csv(d_fit, out("fitted_monthly.csv"))

d_aem_bands |>
  mutate(r2adj = round(100 * r2adj, 1), band = factor(band, BANDS)) |>
  pivot_wider(names_from = band, values_from = r2adj, names_sort = TRUE) |>
  left_join(d_aem_tot |> mutate(`total (%)` = round(100 * r2adj, 1)) |> select(nms, `total (%)`, `selected` = n_selected), by = "nms") |>
  mutate(nms = factor(nms, NMS_ORD)) |>
  arrange(nms) |>
  kable(caption = "Adjusted R2 (%) of the eigenfunction model by time-scale band (full-cycle period: sub-annual <= 1.5 yr, interannual 1.5-8, decadal 8-15, trend-like > 15)")
```

```{r}
#| label: fig_aem_bands
#| fig-width: 9
#| fig-height: 4.8

p <- d_aem_bands |>
  mutate(nms = factor(nms, rev(NMS_ORD)), band = factor(band, BANDS)) |>
  ggplot(aes(r2adj, nms)) +
  geom_col(aes(fill = band), width = 0.62, color = PAL$surface, linewidth = 0.6, position = position_stack(reverse = TRUE)) +
  scale_fill_manual(values = setNames(PAL_BAND, BANDS), name = "time-scale band") +
  scale_x_continuous(labels = label_percent()) +
  labs(title = "Eigenfunction model: adjusted R2 by time-scale band", x = "adjusted R2", y = NULL) +
  theme_chk()
save_fig(p, "diag_aem_bands.png", 9, 4.8)
```

## 6. A common state space for all sanctuaries

One uncentred, unscaled PCA of the annual mean anomalies of all sanctuaries pooled: the origin is
"at baseline" and units are baseline SDs. Axis 1 is oriented so `sst` loads positive, axis 2 so
`chl` loads positive.

```{r}
#| label: pca

ann_z <- dz |>
  mutate(year = year(date)) |>
  filter(var %in% VARS) |>
  semi_join(imap_dfr(Z_list, \(zi, n) tibble(nms = n, var = zi$vars)), by = c("nms", "var")) |>
  group_by(nms, year, var) |>
  summarize(z = mean(z, na.rm = TRUE)) |>
  ungroup() |>
  pivot_wider(names_from = var, values_from = z) |>
  right_join(expand_grid(nms = NMS_ORD, year = YEARS), by = c("nms", "year"))
for (v in setdiff(VARS, names(ann_z))) ann_z[[v]] <- 0
ann_z <- ann_z |> mutate(across(all_of(VARS), \(x) replace_na(x, 0))) |> arrange(factor(nms, NMS_ORD), year)
X_ann <- as.matrix(ann_z[, VARS])

pc  <- prcomp(X_ann, center = FALSE, scale. = FALSE)
rot <- pc$rotation[, 1:3]
if (rot["sst", 1] < 0) rot[, 1] <- -rot[, 1]
if (rot["chl", 2] < 0) rot[, 2] <- -rot[, 2]
var_share <- pc$sdev^2 / sum(pc$sdev^2)
colnames(rot) <- c("pc1", "pc2", "pc3")

d_load <- bind_rows(
  as_tibble(rot, rownames = "var"),
  tibble(var = "variance_share", pc1 = var_share[1], pc2 = var_share[2], pc3 = var_share[3]))
write_csv(d_load, out("pca_loadings.csv"))

S_ann <- X_ann %*% rot
d_pca_ann <- ann_z |> select(nms, year) |> bind_cols(as_tibble(S_ann))
write_csv(d_pca_ann, out("pca_scores_annual.csv"))

d_pca_fit <- d_fit |>
  pivot_wider(names_from = var, values_from = z_fit)
for (v in setdiff(VARS, names(d_pca_fit))) d_pca_fit[[v]] <- 0
d_pca_fit <- d_pca_fit |> mutate(across(all_of(VARS), \(x) replace_na(x, 0)))
S_fit <- as.matrix(d_pca_fit[, VARS]) %*% rot[, 1:2]
d_pca_fit <- d_pca_fit |> select(nms, date) |> bind_cols(as_tibble(S_fit))
write_csv(d_pca_fit, out("pca_scores_fit.csv"))

# 95% prediction ellipse of the ten baseline annual means ----
ellipse_of <- function(S, n_pts = 100) {
  n  <- nrow(S)
  r2 <- 2 * (n - 1) * (n + 1) / (n * (n - 2)) * qf(0.95, 2, n - 2)
  L  <- t(chol(cov(S)))
  th <- seq(0, 2 * pi, length.out = n_pts + 1)
  t(colMeans(S) + sqrt(r2) * L %*% rbind(cos(th), sin(th)))
}
d_ell <- d_pca_ann |>
  filter(year <= 2012) |>
  group_by(nms) |>
  group_modify(\(d, k) { e <- ellipse_of(as.matrix(d[, c("pc1", "pc2")])); tibble(pc1 = e[, 1], pc2 = e[, 2]) }) |>
  ungroup() |>
  arrange(factor(nms, NMS_ORD))
write_csv(d_ell, out("pca_baseline_ellipse.csv"))

d_load |> mutate(across(-var, \(x) round(x, 2))) |>
  kable(caption = "Loadings of the common state space (last row: share of the total sum of squares)")
```

```{r}
#| label: fig_pca
#| fig-width: 10
#| fig-height: 8

p <- d_pca_ann |>
  mutate(nms = factor(nms, NMS_ORD), post = year >= 2013) |>
  ggplot(aes(pc1, pc2)) +
  geom_hline(yintercept = 0, color = PAL$axis, linewidth = 0.2) +
  geom_vline(xintercept = 0, color = PAL$axis, linewidth = 0.2) +
  geom_polygon(data = d_ell |> mutate(nms = factor(nms, NMS_ORD)), fill = PAL$grid, color = PAL$axis, linewidth = 0.3) +
  geom_path(color = PAL$axis, linewidth = 0.3) +
  geom_point(aes(color = year), size = 1.2) +
  facet_wrap(~nms, ncol = 4) +
  scale_color_gradient(low = PAL$blue, high = PAL$orange) +
  labs(
    title = "Annual trajectories in the common state space",
    subtitle = glue("PC1 {percent(var_share[1], 1)}, PC2 {percent(var_share[2], 1)}; grey = 95% prediction ellipse of the 2003-2012 annual means"),
    x = "PC1", y = "PC2") +
  theme_chk()
save_fig(p, "diag_pca.png", 10, 8)
```

## 7. Documented events

Each catalogued event is scored on its own months (no padding) by its **prominence**: the mean
`sigma_cv` over the event months minus the mean over the 24 months before it (at least 6 needed).
The observed prominence is ranked among the prominences of every window of the same length in that
sanctuary's series (`pct_prom`, 0-100); the event registers when `pct_prom >= 90`. This compares an
event with its own sanctuary's background, unlike "any month beyond 2 sigma" (`exceed_any`, kept),
which chance alone satisfies for most long windows once the background rate is high. The peak,
signature and expected-direction check use the event months plus one month either side, because
catalogue dates are approximate. Group test: each event is moved to a random start month (same
sanctuary, same length) in each of `r N_EV_SUR` draws.

```{r}
#| label: events

N_BEFORE     <- 24
N_BEFORE_MIN <- 6
PCT_REG      <- 90

d_ev <- read_csv(events_csv, col_types = cols(.default = "c")) |>
  mutate(
    date_start = as_date(date_start),
    date_end   = as_date(coalesce(date_end, as.character(date_start))),
    resolvable = as.logical(toupper(str_trim(resolvable))))
stopifnot(all(c("event_id", "nms", "date_start", "date_end", "vars_expected", "resolvable") %in% names(d_ev)))

# event months, and event months padded by one month for the signature
ev_core <- map2(d_ev$date_start, d_ev$date_end, \(d0, d1)
  which(WIN_DATES >= floor_date(d0, "month") & WIN_DATES <= floor_date(d1, "month")))
ev_idx  <- map2(d_ev$date_start, d_ev$date_end, \(d0, d1) {
  w0 <- floor_date(d0, "month") %m-% months(1)
  w1 <- floor_date(d1, "month") %m+% months(1)
  which(WIN_DATES >= w0 & WIN_DATES <= w1)
})
ex_list <- map(set_names(NMS_ORD), \(n) replace_na(d_dep$exceed[d_dep$nms == n], FALSE))
sc_list <- map(set_names(NMS_ORD), \(n) d_dep$sigma_cv[d_dep$nms == n])

# prominence of the window starting at st (length L) for every valid start ----
prom_all <- function(sc, L) {
  n  <- length(sc)
  cs <- c(0, cumsum(sc))
  if (L < 1 || n - L + 1 < N_BEFORE_MIN + 1) return(tibble(st = integer(), prom = numeric()))
  st <- seq(N_BEFORE_MIN + 1, n - L + 1)
  b0 <- pmax(1, st - N_BEFORE)
  tibble(
    st   = st,
    prom = (cs[st + L] - cs[st]) / L - (cs[st] - cs[b0]) / (st - b0))
}
# check against a direct computation
sc_t <- sc_list[[1]]; pa <- prom_all(sc_t, 5)
stopifnot(abs(pa$prom[pa$st == 40] - (mean(sc_t[40:44]) - mean(sc_t[16:39]))) < 1e-12,
          abs(pa$prom[1] - (mean(sc_t[7:11]) - mean(sc_t[1:6]))) < 1e-12)

parse_expected <- function(s) {
  if (is.na(s) || str_trim(s) == "") return(tibble(var = character(), sgn = numeric()))
  tok <- str_trim(str_split(s, ";")[[1]])
  tok <- tok[str_detect(tok, "^[a-z_]+[+-]$")]
  tibble(var = str_remove(tok, "[+-]$"), sgn = if_else(str_ends(tok, "\\+"), 1, -1))
}
score_event <- function(nms_i, core, idx, vars_expected) {
  na_z <- set_names(rep(NA_real_, length(VARS)), VARS)
  res  <- list(sigma_max = NA_real_, date_peak = as_date(NA), pct_rank = NA_real_, exceed_any = NA,
               direction_ok = NA, sigma_event = NA_real_, sigma_before = NA_real_, prominence = NA_real_,
               pct_prom = NA_real_, null_pct = numeric(), z = na_z, zx = na_z)
  if (!(nms_i %in% NMS_ORD) || length(idx) == 0) return(res)
  sc  <- sc_list[[nms_i]]
  if (all(is.na(sc[idx]))) return(res)
  j   <- idx[which.max(sc[idx])]
  Z   <- Z_list[[nms_i]]$Z
  res$sigma_max  <- sc[j]
  res$date_peak  <- WIN_DATES[j]
  res$pct_rank   <- 100 * rank(sc, na.last = "keep")[j] / sum(!is.na(sc))
  res$exceed_any <- any(ex_list[[nms_i]][idx])
  for (v in colnames(Z)) {
    res$z[v]  <- Z[j, v]
    w <- Z[idx, v]
    if (any(!is.na(w))) res$zx[v] <- w[which.max(abs(w))]
  }
  ex <- parse_expected(vars_expected) |> filter(var %in% colnames(Z))
  if (nrow(ex) > 0)
    res$direction_ok <- all(map2_lgl(ex$var, ex$sgn, \(v, s) any(s * Z[idx, v] >= 1.5, na.rm = TRUE)))
  # prominence on the event months
  if (length(core) > 0) {
    st0 <- core[1]; L <- length(core)
    res$sigma_event <- mean(sc[core])
    if (st0 - 1 >= N_BEFORE_MIN) {
      res$sigma_before <- mean(sc[max(1, st0 - N_BEFORE):(st0 - 1)])
      res$prominence   <- res$sigma_event - res$sigma_before
      np <- prom_all(sc, L)$prom
      if (length(np) > 0) {
        res$pct_prom <- 100 * mean(np <= res$prominence + 1e-12)
        res$null_pct <- 100 * ecdf(np)(np)       # percentile of every possible window, for the group test
      }
    }
  }
  res
}
ev_scores <- pmap(list(d_ev$nms, ev_core, ev_idx, d_ev$vars_expected), score_event)
d_ev_scored <- d_ev |>
  mutate(
    n_months     = map_int(ev_core, length),
    sigma_max    = map_dbl(ev_scores, "sigma_max"),
    date_peak    = do.call(c, map(ev_scores, "date_peak")),
    pct_rank     = map_dbl(ev_scores, "pct_rank"),
    exceed_any   = map_lgl(ev_scores, "exceed_any"),
    direction_ok = map_lgl(ev_scores, "direction_ok"),
    sigma_event  = map_dbl(ev_scores, "sigma_event"),
    sigma_before = map_dbl(ev_scores, "sigma_before"),
    prominence   = map_dbl(ev_scores, "prominence"),
    pct_prom     = map_dbl(ev_scores, "pct_prom"),
    registered   = pct_prom >= PCT_REG) |>
  bind_cols(
    map_dfr(ev_scores, \(s) as_tibble(as.list(s$z))) |> rename_with(\(x) paste0("z_", x)),
    map_dfr(ev_scores, \(s) as_tibble(as.list(s$zx))) |> rename_with(\(x) paste0("zx_", x)))
write_csv(d_ev_scored, out("events_scored.csv"))

# group test: random start month per event (same sanctuary, same length) ----
ok_ev  <- which(!is.na(d_ev_scored$registered) & !is.na(d_ev_scored$resolvable))
res_ok <- d_ev_scored$resolvable[ok_ev]
obs_reg <- d_ev_scored$registered[ok_ev]
set.seed(4242)
null_reg <- t(replicate(N_EV_SUR, {
  h <- map_lgl(ok_ev, \(i) { np <- ev_scores[[i]]$null_pct; np[sample.int(length(np), 1)] >= PCT_REG })
  c(`TRUE` = mean(h[res_ok]), `FALSE` = mean(h[!res_ok]))
}))
d_ev_test <- map_dfr(c(TRUE, FALSE), \(g) {
  n <- sum(res_ok == g)
  if (n == 0) return(tibble(group = if (g) "resolvable" else "not resolvable", n = 0L,
                            hit_rate = NA_real_, null_mean = NA_real_, null_q95 = NA_real_, p_value = NA_real_))
  nh <- null_reg[, as.character(g)]
  ob <- mean(obs_reg[res_ok == g])
  tibble(group = if (g) "resolvable" else "not resolvable", n = n, hit_rate = ob,
         null_mean = mean(nh), null_q95 = unname(quantile(nh, 0.95)), p_value = p_emp(nh, ob))
})
write_csv(d_ev_test, out("events_test.csv"))

n_unscored <- nrow(d_ev_scored) - length(ok_ev)
if (n_unscored > 0) note("{n_unscored} catalogue events not scored (outside the window, fewer than {N_BEFORE_MIN} prior months, sanctuary absent, or resolvable missing)")

d_ev_test |> mutate(across(c(hit_rate, null_mean, null_q95), \(x) round(x, 2)), p_value = round(p_value, 3)) |>
  kable(caption = glue("Share of events registered (prominence at or above the {PCT_REG}th percentile of same-length windows) vs {N_EV_SUR} random placements"))
d_ev_scored |>
  filter(!is.na(pct_prom)) |>
  mutate(nms = factor(nms, NMS_ORD)) |>
  arrange(desc(resolvable), nms, date_start) |>
  transmute(
    event_id, resolvable, months = n_months, sigma_event = round(sigma_event, 1),
    before = round(sigma_before, 1), pct_prom = round(pct_prom), registered, exceed_any, direction_ok,
    signature = pmap_chr(across(paste0("zx_", VARS)), \(...) {
      v <- c(...); v <- v[!is.na(v) & abs(v) >= 1.5]
      paste(sprintf("%s %+.1f", str_remove(names(v), "zx_"), v), collapse = ", ")
    })) |>
  kable(caption = "Events scored: mean sigma in the event months and the 24 months before, prominence percentile, and signature (variables reaching |z| >= 1.5 in the padded window)")
```

```{r}
#| label: mhw_check

has_mhw <- file.exists(mhw_csv)
if (has_mhw) {
  d_mhw <- read_csv(mhw_csv) |> mutate(date = as_date(date))
  d_mhw_chk <- d_mhw |>
    inner_join(dz |> filter(var == "sst") |> select(nms, date, z_sst = z), by = c("nms", "date")) |>
    filter(!is.na(z_sst)) |>
    group_by(nms) |>
    summarize(
      n_months          = n(),
      rho_mhw_zsst      = suppressWarnings(cor(mhw_days, z_sst, method = "spearman", use = "complete.obs")),
      rho_mcs_negzsst   = suppressWarnings(cor(mcs_days, -z_sst, method = "spearman", use = "complete.obs")),
      n_mhw15           = sum(mhw_days >= 15, na.rm = TRUE),
      share_mhw15_zsst1 = if (n_mhw15 > 0) mean(z_sst[mhw_days >= 15 & !is.na(mhw_days)] >= 1) else NA_real_) |>
    mutate(nms = factor(nms, NMS_ORD)) |>
    arrange(nms) |>
    mutate(nms = as.character(nms))
  write_csv(d_mhw_chk, out("mhw_check.csv"))
  d_mhw_chk |> mutate(across(where(is.double), \(x) round(x, 2))) |>
    kable(caption = "Daily-SST cross-check: marine heatwave / cold-spell days (OISST) vs monthly z of sst")
} else {
  note("data/events/mhw_monthly.csv is absent: daily-SST cross-check skipped.")
}
```

## 8. Basin-scale climate modes

Each index (standardized over the window) is lagged 0-12 months; the lag with the largest adjusted
R2 of an RDA of the anomalies is kept, and tested against phase-randomized surrogates of the index
(lag re-optimized in each). A joint model with all indices at their best lags is partitioned against
the selected eigenfunctions, and the departure is recomputed with the mode-related part removed.

```{r}
#| label: modes

has_idx <- file.exists(indices_csv)
d_mode_res <- NULL
if (has_idx) {
  d_idx <- read_csv(indices_csv) |>
    mutate(date = as_date(date), index = tolower(index)) |>
    filter(index %in% INDICES, !is.na(value)) |>
    distinct(index, date, .keep_all = TRUE)
  idx_present <- intersect(INDICES, unique(d_idx$index))
  has_idx <- length(idx_present) > 0
}
if (!has_idx) {
  note("data/context/indices_monthly.csv is absent (or has none of {paste(INDICES, collapse = ', ')}): climate-mode section skipped.")
} else {
  IDX_DATES <- seq(WIN_BEG %m-% months(max(LAGS)), WIN_END, by = "month")
  idx_mat <- sapply(idx_present, \(i) {
    s <- d_idx |> filter(index == i)
    v <- s$value[match(IDX_DATES, s$date)]
    w <- IDX_DATES >= WIN_BEG
    (v - mean(v[w], na.rm = TRUE)) / sd(v[w], na.rm = TRUE)
  })
  lagged <- \(v, L) v[(max(LAGS) + 1 - L):(length(v) - L)]
  r2_by_lag <- \(Y, v) map_dbl(LAGS, \(L) {
    x <- lagged(v, L)
    if (mean(!is.na(x)) < 1 - MAX_NA) NA_real_ else r2adj_fast(Y, matrix(x))
  })
  tibble(index = idx_present,
         first = map_chr(idx_present, \(i) format(min(d_idx$date[d_idx$index == i]), "%Y-%m")),
         last  = map_chr(idx_present, \(i) format(max(d_idx$date[d_idx$index == i]), "%Y-%m")),
         `missing months in window (lag 0)` = map_int(idx_present, \(i) sum(is.na(lagged(idx_mat[, i], 0))))) |>
    kable(caption = "Climate indices used")

  # indices assigned a priori by basin
  idx_of <- \(n) intersect(MODE_SETS[[BASIN[[n]]]], idx_present)
  pairs  <- map_dfr(NMS_ORD, \(n) tibble(nms = n, index = idx_of(n)))
  tibble(nms = NMS_ORD, basin = BASIN[NMS_ORD], indices = map_chr(NMS_ORD, \(n) paste(idx_of(n), collapse = ", "))) |>
    kable(caption = glue("Indices assigned by basin; lags 0-{max(LAGS)} months (index leading)"))

  modes_single_csv <- out("modes_single.csv")
  modes_ok <- is_fresh(modes_single_csv, c(state_csv, indices_csv)) && {
    m <- read_csv(modes_single_csv)
    nrow(m) == nrow(pairs) && all(paste(m$nms, m$index) %in% paste(pairs$nms, pairs$index)) &&
      all(m$lag_best <= max(LAGS)) }
  if (!modes_ok) {
    set.seed(777)
    sur <- map(set_names(idx_present), \(i) replicate(N_MODE_SUR, phase_rand(idx_mat[, i])))
    res <- mclapply(NMS_ORD, \(n) {
      Y <- na0(Z_list[[n]]$Z)
      map_dfr(idx_of(n), \(i) {
        r2  <- r2_by_lag(Y, idx_mat[, i])
        Lb  <- LAGS[which.max(r2)]
        xb  <- lagged(idx_mat[, i], Lb)
        mx  <- apply(sur[[i]], 2, \(v) max(r2_by_lag(Y, v), na.rm = TRUE))
        rr  <- \(v) if (v %in% colnames(Y)) cor(xb, Y[, v], use = "complete.obs") else NA_real_
        tibble(nms = n, index = i, lag_best = Lb, r2adj = max(r2, na.rm = TRUE),
               r_sst = rr("sst"), r_sss = rr("sss"), r_chl = rr("chl"), p_value = p_emp(mx, max(r2, na.rm = TRUE)))
      })
    }, mc.cores = N_CORE)
    stopifnot(!any(map_lgl(res, \(r) inherits(r, "try-error"))))
    write_csv(bind_rows(res), modes_single_csv)
  }
  d_modes_single <- read_csv(modes_single_csv)

  # joint model, variance partitioning, per-variable r2, residual departure ----
  d_mode_res <- map(set_names(NMS_ORD), \(n) {
    zi <- Z_list[[n]]
    Y  <- na0(zi$Z)
    lb <- d_modes_single |> filter(nms == n)
    Xj <- sapply(set_names(lb$index), \(i) lagged(idx_mat[, i], lb$lag_best[lb$index == i]))
    n_fill <- sum(is.na(Xj))
    Xj[is.na(Xj)] <- 0
    Xc <- sweep(Xj, 2, colMeans(Xj))
    r2_modes <- r2adj_fast(Y, Xc)
    ks <- d_aem_sel$k[d_aem_sel$nms == n]
    vp <- if (length(ks) > 0) varpart(Y, Xc, E_aem[, paste0("AEM", ks), drop = FALSE])$part$indfract$Adj.R.square else rep(NA_real_, 4)
    B  <- qr.coef(qr(cbind(1, Xc)), Y)
    Zr <- zi$Z - Xc %*% B[-1, , drop = FALSE]
    st <- dep_stats(Zr)
    list(
      # vegan's indfract order: [a] = X1|X2, [b] = X2|X1, [c] = shared, [d] = residuals
      total = tibble(nms = n, r2adj_modes = r2_modes, modes_unique = vp[1], shared = vp[3],
                     aem_unique = vp[2], residual = vp[4],
                     lags = paste(glue("{lb$index}@{lb$lag_best}"), collapse = ";"), n_filled = n_fill),
      by_var = tibble(nms = n, var = zi$vars, r2adj = map_dbl(zi$vars, \(v) r2adj_fast(Y[, v, drop = FALSE], Xc))),
      sigma_res = st$sigma_cv,
      e_res = st$e_per)
  })
  d_modes_total <- map_dfr(d_mode_res, "total")
  stopifnot(with(d_modes_total, all(is.na(shared) | abs(modes_unique + shared - r2adj_modes) < 1e-6)))
  d_modes_var   <- map_dfr(d_mode_res, "by_var")
  write_csv(d_modes_total, out("modes_total.csv"))
  write_csv(d_modes_var, out("modes_by_var.csv"))

  d_modes_single |>
    mutate(cell = glue("{round(100 * r2adj)}{if_else(p_value <= 0.05, '*', '')} (lag {lag_best})")) |>
    select(nms, index, cell) |>
    pivot_wider(names_from = index, values_from = cell) |>
    left_join(d_modes_total |> transmute(nms, `joint (%)` = round(100 * r2adj_modes),
                                         `modes only` = round(100 * modes_unique), shared = round(100 * shared),
                                         `eigenfunctions only` = round(100 * aem_unique)), by = "nms") |>
    mutate(nms = factor(nms, NMS_ORD)) |>
    arrange(nms) |>
    kable(caption = "Adjusted R2 (%) of each index at its best lag (* p <= 0.05, phase-randomized surrogates), joint model and its partition with the eigenfunctions")
  d_modes_var |>
    mutate(r2adj = round(100 * r2adj)) |>
    pivot_wider(names_from = var, values_from = r2adj) |>
    mutate(nms = factor(nms, NMS_ORD)) |>
    arrange(nms) |>
    kable(caption = "Adjusted R2 (%) of each variable's z on all indices at their best lags")
}
```

```{r}
#| label: write_departure

# departure and period exceedance, with the mode-removed versions when available ----
d_dep_out <- d_dep |>
  mutate(sigma_res = if (!is.null(d_mode_res)) unlist(map(d_mode_res[NMS_ORD], "sigma_res")) else NA_real_) |>
  select(nms, date, sigma, sigma_cv, exceed, d_rms, sigma_run12, exceed_run12, sigma_res)
write_csv(d_dep_out, out("departure_monthly.csv"))

d_eper_out <- d_eper |>
  left_join(
    if (!is.null(d_mode_res))
      imap_dfr(d_mode_res, \(r, n) tibble(nms = n, period = PERIODS$period, e_res_pct = unname(r$e_res)))
    else tibble(nms = character(), period = character(), e_res_pct = numeric()),
    by = c("nms", "period")) |>
  select(nms, period, e_pct, sigma_median, null_q50, null_q95, p_value, e_res_pct)
write_csv(d_eper_out, out("exceedance_period.csv"))

if (!is.null(d_mode_res)) {
  d_eper_out |>
    filter(period %in% c("2016-2020", "2021-2025")) |>
    transmute(nms, period, cell = glue("{round(e_pct)} -> {round(e_res_pct)} ({round(null_q95)})")) |>
    pivot_wider(names_from = period, values_from = cell) |>
    mutate(nms = factor(nms, NMS_ORD)) |>
    arrange(nms) |>
    kable(caption = "Exceedance (%) before -> after removing the climate-mode part (null 95th percentile of the main analysis in brackets)")
}
```

## 9. Dated shifts (supporting)

E-divisive change points on the annual mean anomalies (first year of each new segment). Known
product changes to keep in mind when reading them: GLORYS12 atmospheric forcing change in 2019 and
the switch to the interim stream in 2021; ocean-color sensor changes in the merged chlorophyll
record in 2012 (VIIRS), 2016 and 2018 (OLCI-A, OLCI-B) and 2023. Nearly every year from 2011 to 2024
is within a year of one of these, so no per-shift flag is given.

```{r}
#| label: shifts

shifts_csv <- out("shifts.csv")
if (!is_fresh(shifts_csv) || !identical(names(read_csv(shifts_csv, n_max = 1)), c("nms", "year")) ||
    !all(read_csv(shifts_csv)$nms %in% NMS_ORD)) {
  res <- mclapply(NMS_ORD, \(n) {
    set.seed(seed_of(n, 2))
    X  <- as.matrix(ann_z[ann_z$nms == n, Z_list[[n]]$vars])
    yr <- ann_z$year[ann_z$nms == n]
    i  <- e.divisive(X, sig.lvl = 0.05, R = 499, min.size = 3)$estimates
    i  <- i[i > 1 & i <= nrow(X)]
    tibble(nms = rep(n, length(i)), year = as.integer(yr[i]))
  }, mc.cores = N_CORE)
  stopifnot(!any(map_lgl(res, \(r) inherits(r, "try-error"))))
  bind_rows(res) |> write_csv(shifts_csv)
}
d_shifts <- read_csv(shifts_csv, col_types = "ci")
tibble(nms = NMS_ORD) |>
  left_join(d_shifts |> group_by(nms) |>
              summarize(`shift years (first year of new segment)` = paste(year, collapse = ", ")),
            by = "nms") |>
  kable(caption = "E-divisive shifts on annual means")
```

## 10. Sensitivity

The full scoring (standardization, envelope, exceedance) is rerun under alternative choices.

```{r}
#| label: sensitivity

variant_e <- function(d_long, vars, base_beg = BASE_BEG, base_end = BASE_END, thr = SIGMA_THR, win_beg = WIN_BEG) {
  dates <- seq(win_beg, WIN_END, by = "month")
  yr    <- year(dates)
  is_b  <- dates >= base_beg & dates <= base_end
  dzv   <- standardize(d_long |> filter(var %in% vars), base_beg, base_end) |>
    filter(date >= win_beg, date <= WIN_END)
  map_dfr(NMS_ORD, \(n) {
    zi <- build_Z(dzv, n, vars, is_b, dates)
    if (ncol(zi$Z) < 2) return(NULL)
    ex <- score_departure(zi$Z, is_b, yr)$sigma_cv > thr
    tibble(
      nms         = n,
      e_2016_2020 = 100 * mean(ex[between(yr, 2016, 2020)], na.rm = TRUE),
      e_2021_2025 = 100 * mean(ex[between(yr, 2021, 2025)], na.rm = TRUE))
  })
}
read_occci <- \() read_csv(occci_csv) |> mutate(date = as_date(date)) |> select(nms, date, chl_occci)
swap_var <- \(d, from, to) d |> filter(var != to) |> mutate(var = if_else(var == from, to, var))

variants <- list(
  main               = \() variant_e(d_state, VARS),
  baseline_2006_2015 = \() variant_e(d_state, VARS, as_date("2006-01-01"), as_date("2015-12-01")),
  thr3               = \() variant_e(d_state, VARS, thr = 3),
  with_pp            = \() if ("pp" %in% d_state$var) variant_e(d_state, c(VARS, "pp")) else NULL,
  unmasked           = \() if (all(c("sbt_all", "chl_all") %in% d_state$var))
    variant_e(d_state |> swap_var("sbt_all", "sbt") |> swap_var("chl_all", "chl"), VARS) else NULL,
  no_chl             = \() variant_e(d_state, setdiff(VARS, "chl")),
  occci              = \() {
    if (!file.exists(occci_csv)) return(NULL)
    d_state |>
      left_join(read_occci(), by = c("nms", "date")) |>
      mutate(value = if_else(var == "chl", if_else(!is.na(value) & !is.na(chl_occci), chl_occci, NA_real_), value)) |>
      select(-chl_occci) |>
      variant_e(VARS)
  },
  # 1998-01..2025-09 with a 1998-2012 baseline; chlorophyll from oc-cci (the only record spanning it consistently)
  long_baseline      = \() {
    if (!file.exists(occci_csv)) return(NULL)
    d_state |>
      filter(var != "chl") |>
      bind_rows(read_occci() |> transmute(nms, date, var = "chl", value = chl_occci)) |>
      variant_e(VARS, as_date("1998-01-01"), as_date("2012-12-01"), win_beg = as_date("1998-01-01"))
  })
d_sens <- imap_dfr(variants, \(f, v) { r <- f(); if (is.null(r)) NULL else mutate(r, variant = v) }) |>
  select(nms, variant, e_2016_2020, e_2021_2025)
missing_variants <- setdiff(names(variants), unique(d_sens$variant))
if (length(missing_variants) > 0) note("sensitivity variants skipped (input absent): {paste(missing_variants, collapse = ', ')}")

# the main variant must reproduce section 4 exactly
chk_main <- d_sens |> filter(variant == "main") |>
  left_join(d_eper |> filter(period == "2021-2025") |> select(nms, e_pct), by = "nms")
stopifnot(max(abs(chk_main$e_2021_2025 - chk_main$e_pct)) < 1e-9)
write_csv(d_sens, out("sensitivity.csv"))

d_season <- d_dep |>
  mutate(
    yr = year(date), mo = month(date),
    season = case_when(mo %in% c(12, 1, 2) ~ "DJF", mo %in% 3:5 ~ "MAM", mo %in% 6:8 ~ "JJA", TRUE ~ "SON")) |>
  cross_join(PERIODS) |>
  filter(yr >= y0, yr <= y1) |>
  group_by(nms, season, period) |>
  summarize(e_pct = 100 * mean(exceed, na.rm = TRUE)) |>
  ungroup() |>
  arrange(factor(nms, NMS_ORD), factor(season, SEASONS), factor(period, PERIODS$period))
write_csv(d_season, out("exceedance_season.csv"))

d_sens |>
  mutate(cell = glue("{round(e_2016_2020)} / {round(e_2021_2025)}")) |>
  select(nms, variant, cell) |>
  pivot_wider(names_from = variant, values_from = cell) |>
  mutate(nms = factor(nms, NMS_ORD)) |>
  arrange(nms) |>
  kable(caption = "Exceedance (%) 2016-2020 / 2021-2025 under each variant")
d_season |>
  filter(period == "2021-2025") |>
  mutate(e_pct = round(e_pct)) |>
  pivot_wider(names_from = season, values_from = e_pct) |>
  select(-period) |>
  mutate(nms = factor(nms, NMS_ORD)) |>
  arrange(nms) |>
  kable(caption = "Exceedance (%) in 2021-2025 by season")
```

### Red-noise reference for the eigenfunction model

The same selection pipeline on `r N_AEM_NULL` red-noise series per sanctuary (VAR(1) of section 3,
`nperm = ` `r NPERM_AEM_NULL`), to show how much adjusted R2, and how much "trend-like" R2,
persistence alone produces. The specification asks for 99 series; one selection on a persistent
series takes 10-30 s, so 99 x 13 would take about 45 minutes and the reference was cut to
`r N_AEM_NULL` series per sanctuary (the "95th percentile" is then close to the maximum).

```{r}
#| label: aem_null

aem_null_csv <- out("aem_null.csv")
if (!is_fresh(aem_null_csv)) {
  n_blk <- 3
  jobs  <- expand_grid(nms = NMS_ORD, blk = seq_len(n_blk))
  res <- mclapply(seq_len(nrow(jobs)), \(j) {
    n   <- jobs$nms[j]
    set.seed(seed_of(n, 100 + jobs$blk[j]))
    reps <- seq(jobs$blk[j], N_AEM_NULL, by = n_blk)
    map_dfr(reps, \(r) {
      sel <- fit_aem_sel(restd(sim_var1(var1_fits[[n]])), E_aem, nperm = NPERM_AEM_NULL)
      tibble(nms = n, rep = r, total = sum(sel$r2adj_increment),
             trend = sum(sel$r2adj_increment[band_of(period_of_k(sel$k)) == "trend-like"]))
    })
  }, mc.cores = N_CORE)
  stopifnot(!any(map_lgl(res, \(r) inherits(r, "try-error"))))
  d_an <- bind_rows(res)
  stopifnot(nrow(d_an) == N_AEM_NULL * length(NMS_ORD))
  d_an |>
    pivot_longer(c(total, trend), names_to = "stat") |>
    mutate(stat = recode(stat, total = "r2adj_total", trend = "r2adj_trend")) |>
    group_by(nms, stat) |>
    summarize(q50 = quantile(value, 0.5), q95 = quantile(value, 0.95)) |>
    ungroup() |>
    arrange(factor(nms, NMS_ORD)) |>
    write_csv(aem_null_csv)
}
d_aem_null <- read_csv(aem_null_csv)

d_aem_tot |>
  select(nms, total = r2adj) |>
  left_join(d_aem_bands |> filter(band == "trend-like") |> select(nms, trend = r2adj), by = "nms") |>
  left_join(d_aem_null |> mutate(q = glue("{round(100 * q50)} / {round(100 * q95)}")) |>
              select(nms, stat, q) |> pivot_wider(names_from = stat, values_from = q), by = "nms") |>
  mutate(total = round(100 * total), trend = round(100 * trend)) |>
  select(nms, `total (%)` = total, `null total, median / 95th` = r2adj_total,
         `trend-like (%)` = trend, `null trend-like, median / 95th` = r2adj_trend) |>
  kable(caption = "Observed eigenfunction-model adjusted R2 vs red noise with the same persistence")
```

## 11. Scorecard

```{r}
#| label: scorecard

wide_var <- function(d, col, prefix) {
  w <- d |> select(nms, var, value = all_of(col)) |>
    pivot_wider(names_from = var, values_from = value, names_prefix = prefix)
  for (v in paste0(prefix, VARS)) if (!v %in% names(w)) w[[v]] <- NA
  w |> select(nms, all_of(paste0(prefix, VARS)))
}
sh25 <- d_shift |> filter(period == "2021-2025")
ep   <- \(p, col = "e_pct") d_eper_out |> filter(period == p) |> select(nms, all_of(col)) |> deframe()

d_score <- NMS_LUT |>
  mutate(
    p                  = P_OF[nms],
    e_base_cv          = ep("base")[nms],
    e_2013_2015        = ep("2013-2015")[nms],
    e_2016_2020        = ep("2016-2020")[nms],
    e_2021_2025        = ep("2021-2025")[nms],
    null_q95_2021_2025 = ep("2021-2025", "null_q95")[nms],
    p_2021_2025        = ep("2021-2025", "p_value")[nms]) |>
  left_join(d_status |> select(nms, status, delta, trend, form, first_period_outside), by = "nms") |>
  left_join(wide_var(sh25, "mean_z", "mean_z_"), by = "nms") |>
  left_join(wide_var(sh25, "sig", "sig_"), by = "nms") |>
  left_join(wide_var(sh25, "sig_mean", "sig_mean_"), by = "nms") |>
  left_join(d_aem_tot |> select(nms, r2adj_aem = r2adj), by = "nms") |>
  left_join(
    d_aem_bands |>
      mutate(band = recode(band, `trend-like` = "r2_trend", decadal = "r2_decadal",
                           interannual = "r2_interannual", `sub-annual` = "r2_subannual")) |>
      pivot_wider(names_from = band, values_from = r2adj),
    by = "nms") |>
  mutate(
    r2adj_modes     = if (exists("d_modes_total")) d_modes_total$r2adj_modes[match(nms, d_modes_total$nms)] else NA_real_,
    e_res_2021_2025 = ep("2021-2025", "e_res_pct")[nms]) |>
  left_join(
    d_ev_scored |>
      group_by(nms) |>
      summarize(
        n_events                = n(),
        n_resolvable            = sum(resolvable %in% TRUE),
        n_registered_resolvable = sum(resolvable %in% TRUE & registered %in% TRUE)),
    by = "nms") |>
  mutate(across(c(n_events, n_resolvable, n_registered_resolvable), \(x) replace_na(x, 0L))) |>
  select(
    nms, name, region, p, e_base_cv, e_2013_2015, e_2016_2020, e_2021_2025, null_q95_2021_2025,
    p_2021_2025, status, delta, trend, form, first_period_outside, all_of(paste0("mean_z_", VARS)),
    all_of(paste0("sig_", VARS)), all_of(paste0("sig_mean_", VARS)), r2adj_aem, r2_trend, r2_decadal, r2_interannual, r2_subannual,
    r2adj_modes, e_res_2021_2025, n_events, n_resolvable, n_registered_resolvable)
write_csv(d_score, out("scorecard.csv"))

top2 <- sh25 |>
  group_by(nms) |>
  slice_max(abs(mean_z), n = 2, with_ties = FALSE) |>
  summarize(`largest shifts` = paste(sprintf("%s %+.1f%s", var, mean_z, if_else(sig_mean, "†", "")), collapse = ", "))
d_score |>
  left_join(top2, by = "nms") |>
  transmute(
    nms, base = round(e_base_cv), `13-15` = round(e_2013_2015), `16-20` = round(e_2016_2020),
    `21-25` = round(e_2021_2025), `null q95` = round(null_q95_2021_2025), status, trend, form,
    `first outside` = first_period_outside, `largest shifts`, `AEM R2` = round(100 * r2adj_aem)) |>
  kable(caption = "Scorecard (exceedance in % of months beyond 2 sigma)")
```

## Checks

```{r}
#| label: checks

schemas <- list(
  z_monthly.csv            = c("nms", "date", "var", "value", "x", "z"),
  z_temp_1993.csv          = c("nms", "date", "var", "value", "x", "z"),
  departure_monthly.csv    = c("nms", "date", "sigma", "sigma_cv", "exceed", "d_rms", "sigma_run12", "exceed_run12", "sigma_res"),
  null_exceedance.csv      = c("nms", "period", "q50", "q95", "q99"),
  null_var.csv             = c("nms", "var", "period", "q95_share", "q025_mean", "q975_mean"),
  exceedance_period.csv    = c("nms", "period", "e_pct", "sigma_median", "null_q50", "null_q95", "p_value", "e_res_pct"),
  exceedance_annual.csv    = c("nms", "year", "e_pct", "sigma_mean", "n_months"),
  shift_by_var.csv         = c("nms", "var", "period", "mean_z", "share_hi", "share_lo", "null_q95_share", "sig",
                               "null_q025_mean", "null_q975_mean", "sig_mean"),
  aem_selected.csv         = c("nms", "k", "period_yr", "band", "r2adj_increment", "p_value"),
  aem_bands.csv            = c("nms", "band", "r2adj"),
  aem_total.csv            = c("nms", "r2adj", "n_selected", "p"),
  fitted_monthly.csv       = c("nms", "date", "var", "z_fit"),
  aem_null.csv             = c("nms", "stat", "q50", "q95"),
  pca_loadings.csv         = c("var", "pc1", "pc2", "pc3"),
  pca_scores_annual.csv    = c("nms", "year", "pc1", "pc2", "pc3"),
  pca_scores_fit.csv       = c("nms", "date", "pc1", "pc2"),
  pca_baseline_ellipse.csv = c("nms", "pc1", "pc2"),
  events_test.csv          = c("group", "n", "hit_rate", "null_mean", "null_q95", "p_value"),
  shifts.csv               = c("nms", "year"),
  sensitivity.csv          = c("nms", "variant", "e_2016_2020", "e_2021_2025"),
  exceedance_season.csv    = c("nms", "season", "period", "e_pct"))
if (has_mhw) schemas$mhw_check.csv <- c("nms", "n_months", "rho_mhw_zsst", "rho_mcs_negzsst", "n_mhw15", "share_mhw15_zsst1")
if (has_idx) {
  schemas$modes_single.csv <- c("nms", "index", "lag_best", "r2adj", "r_sst", "r_sss", "r_chl", "p_value")
  schemas$modes_total.csv  <- c("nms", "r2adj_modes", "modes_unique", "shared", "aem_unique", "residual", "lags", "n_filled")
  schemas$modes_by_var.csv <- c("nms", "var", "r2adj")
}
n_nms <- length(NMS_ORD)
n_var_used <- sum(P_OF)
rows_expected <- list(
  departure_monthly.csv    = n_nms * N_WIN,
  null_exceedance.csv      = n_nms * nrow(PERIODS),
  null_var.csv             = n_var_used * nrow(PERIODS_VAR),
  exceedance_period.csv    = n_nms * nrow(PERIODS),
  exceedance_annual.csv    = n_nms * length(YEARS),
  shift_by_var.csv         = n_var_used * 2,
  aem_bands.csv            = n_nms * length(BANDS),
  aem_total.csv            = n_nms,
  fitted_monthly.csv       = n_var_used * N_WIN,
  aem_null.csv             = n_nms * 2,
  pca_loadings.csv         = length(VARS) + 1,
  pca_scores_annual.csv    = n_nms * length(YEARS),
  pca_scores_fit.csv       = n_nms * N_WIN,
  pca_baseline_ellipse.csv = n_nms * 101,
  events_test.csv          = 2,
  exceedance_season.csv    = n_nms * length(SEASONS) * nrow(PERIODS),
  scorecard.csv            = n_nms)
if (has_idx) rows_expected$modes_total.csv <- n_nms

d_chk <- imap_dfr(schemas, \(cols, f) {
  d <- read_csv(out(f), col_types = cols(.default = "c"))
  tibble(file = f, rows = nrow(d), schema_ok = identical(names(d), cols),
         rows_ok = is.null(rows_expected[[f]]) || nrow(d) == rows_expected[[f]])
})
d_chk <- bind_rows(d_chk, tibble(
  file = "scorecard.csv", rows = nrow(read_csv(out("scorecard.csv"))),
  schema_ok = ncol(read_csv(out("scorecard.csv"))) == 43, rows_ok = rows == n_nms))
kable(d_chk, caption = "Schema and row-count checks of the files written")
stopifnot(all(d_chk$schema_ok), all(d_chk$rows_ok))

# content checks
stopifnot(
  !anyNA(d_dep_out$sigma_cv), !anyNA(d_dep_out$exceed),
  all(d_eper_out$p_value > 0 & d_eper_out$p_value <= 1),
  all(abs(colSums(d_load[1:6, -1]^2) - 1) < 1e-9),
  abs(sum(pc$sdev^2 / sum(pc$sdev^2)) - 1) < 1e-9,
  all(d_score$status %in% c("outside", "within")),
  all(d_score$trend %in% c("increasing", "steady", "decreasing")),
  all(!is.na(d_score$form)),
  nrow(read_csv(out("z_monthly.csv"))) == nrow(dz),
  all(c("exceed_any", "sigma_event", "sigma_before", "prominence", "pct_prom", "registered") %in%
        names(read_csv(out("events_scored.csv")))),
  setequal(d_score$nms, NMS_ORD))
cat(glue("Render time: {round(as.numeric(difftime(Sys.time(), t_start, units = 'mins')), 1)} min"), "\n")
```

## Session info

```{r}
#| label: session

sessionInfo()
```
